---
title: "Advanced R Workbook (Metaprogramming)"
date: '2023-02-19'
output:
  html_notebook:
    theme:
      version: 5
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
---

Workbook for completing quizzes and exercises from the "Metaprogramming" chapters of [Advanced R](https://adv-r.hadley.nz/index.html), second edition, with comparisons to solutions from [Advanced R Solutions](https://advanced-r-solutions.rbind.io).

```{r setup, message = FALSE, warning = FALSE}
library(lobstr)
library(rlang)

# from https://github.com/hadley/adv-r/blob/master/common.R
knitr::opts_chunk$set(
  comment = "#>",
  fig.align = "center"
)

knitr::knit_hooks$set(
  small_mar = function(before, options, envir) {
    if (before) {
      par(mar = c(4.1, 4.1, 0.5, 0.5)) # nolint: undesirable_function_linter.
    }
  }
)
```

# Introduction

This workbook includes answers and solutions to the quizzes and exercises from [Advanced R](https://adv-r.hadley.nz/index.html) and [Advanced R Solutions](https://advanced-r-solutions.rbind.io), organized by chapter. It includes excerpts from both books, copied here.

**WARNING, SPOILERS!** If you haven't read Advanced R and intend to complete the quizzes and exercises, don't read this notebook. It contains my (potentially wrong) answers to both.

# 17 Big picture

Metaprogramming is the hardest topic in this book because it brings together many formerly unrelated topics and forces you grapple with issues that you probably haven't thought about before. You'll also need to learn a lot of new vocabulary, and at first it will seem like every new term is defined by three other terms that you haven't heard of. Even if you're an experienced programmer in another language, your existing skills are unlikely to be much help as few modern popular languages expose the level of metaprogramming that R provides. So don't be surprised if you're frustrated or confused at first; this is a natural part of the process that happens to everyone! 

But I think it's easier to learn metaprogramming now than ever before. Over the last few years, the theory and practice have matured substantially, providing a strong foundation paired with tools that allow you to solve common problems. In this chapter, you'll get the big picture of all the main pieces and how they fit together. 

# 18 Expressions

To compute on the language, we first need to understand its structure. That requires some new vocabulary, some new tools, and some new ways of thinking about R code. The first of these is the distinction between an operation and its result. Take the following code, which multiplies a variable `x` by 10 and saves the result to a new variable called `y`. It doesn't work because we haven't defined a variable called `x`:

```{r, error = TRUE}
rm(x)
try(y <- x * 10)
```

It would be nice if we could capture the intent of the code without executing it. In other words, how can we separate our description of the action from the action itself? 

One way is to use `rlang::expr()`:

```{r}
z <- rlang::expr(y <- x * 10)
z
```

`expr()` returns an expression, an object that captures the structure of the code without evaluating it (i.e. running it). If you have an expression, you can evaluate it with `base::eval()`:

```{r}
x <- 4
eval(z)
y
```

The focus of this chapter is the data structures that underlie expressions. Mastering this knowledge will allow you to inspect and modify captured code, and to generate code with code. We'll come back to `expr()` in Chapter 19, and to `eval()` in Chapter 20.

## 18.2.4 Exercises

1.  Reconstruct the code represented by the trees below:

```{r}
#> █─f
#> └─█─g
#>   └─█─h
#> █─`+`
#> ├─█─`+`
#> │ ├─1
#> │ └─2
#> └─3
#> █─`*`
#> ├─█─`(`
#> │ └─█─`+`
#> │   ├─x
#> │   └─y
#> └─z
```

Answer: code below.

```{r}
lobstr::ast(f(g(h())))
lobstr::ast(1 + 2 + 3)
lobstr::ast((x + y) * z)
```

AR Solutions: Let the source (of the code chunks above) be with you and show you how the ASTs (abstract syntax trees) were produced.

```{r}
ast(f(g(h())))
ast(1 + 2 + 3)
ast((x + y) * z)
```

---

2.  Draw the following trees by hand and then check your answers with
    `lobstr::ast()`.

```{r, eval = FALSE}
f(g(h(i(1, 2, 3))))
f(1, g(2, h(3, i())))
f(g(1, 2), h(3, i(4, 5)))
```

Answer: code below, expression 1:

```{r}
#> █─f
#> └─█─g
#>   └─█─h
#>     └─█─i
#>       └─1
#>       └─2
#>       └─3
lobstr::ast(f(g(h(i(1, 2, 3)))))
```

Expression 2:

```{r}
#> █─f
#> └─1
#> └─█─g
#>   └─2
#>   └─█─h
#>     └─3
#>     └─█─i
lobstr::ast(f(1, g(2, h(3, i()))))
```

Expression 3:

```{r}
#> █─f
#> └─█─g
#> | └─1
#> | └─2
#> └─█─h
#>   └─3
#>   └─█─i
#>     └─4
#>     └─5
lobstr::ast(f(g(1, 2), h(3, i(4, 5))))
```

AR Solutions: Let us delegate the drawing to the `lobstr` package.

---

3.  What's happening with the ASTs below? (Hint: carefully read `?"^"`.)

```{r}
lobstr::ast(`x` + `y`)
lobstr::ast(x ** y) # styler: off
lobstr::ast(1 -> x) # nolint: assignment_linter.
```

Answer: all expressions are changed to their normal forms. In the first expression, the back ticks denoting symbols are not needed. In the third expression, the assignment operator is changed to the standard form. In the second expression, the operator is changed per this note in the R documentation:

> `**` is translated in the parser to `^`, but this was undocumented for many years. It appears as an index entry in Becker *et al* (1988), pointing to the help for `Deprecated` but is not actually mentioned on that page. Even though it had been deprecated in S for 20 years, it was still accepted in R in 2008.

AR Solutions: ASTs start function calls with the name of the function. This is why the call in the first expression is translated into its prefix form. In the second case, `**` is translated by R's parser into `^`. In the last AST, the expression is flipped when R parses it.

---

4.  What is special about the AST below? (Hint: re-read Section
    6.2.1.)

```{r}
lobstr::ast(function(x = 1, y = 2) {}) # nolint: brace_linter.
```

Answer: the AST includes two of the three parts of the function, the `formals()`, `x` and `y`, the `body()`, represented by `{` and `<inline srcref>`.

AR Solutions: The last leaf of the AST is not explicitly specified in the expression. Instead, the `srcref` attribute, which points to the functions source code, is automatically created by base R.

---

5.  What does the call tree of an `if` statement with multiple `else if`
    conditions look like? Why?

Answer: the call tree shows nested if statements; this is because if-else is a function with three arguments, the `cond`, the `cons.expr` and the `alt.expr`. The `alt.expr` for each if-else statement is a new call to if.

```{r}
lobstr::ast(
  if (v == 1) {
    print("one")
  } else if (v == 2) {
    print("two")
  } else if (v == 3) {
    print("three")
  } else if (v == 4) {
    print("four")
  }
)
```

AR Solutions: The AST of nested `else if` statements might look a bit confusing because it contains multiple curly braces. However, we can see that in the `else` part of the AST just another expression is being evaluated, which happens to be an `if` statement and so forth.

```{r}
ast(
  if (FALSE) {
    1
  } else if (FALSE) {
    2
  } else if (TRUE) {
    3
  }
)
```

We can see the structure more clearly if we avoid the curly braces:

```{r}
ast(
  # styler: off
  if (FALSE) 1
  else if (FALSE) 2
  else if (TRUE) 3
  # styler: on
)
```

---

## 18.3.5 Exercises

2.  Which two of the six types of atomic vector can't appear in an expression?
    Why? Similarly, why can't you create an expression that contains an atomic 
    vector of length greater than one?

Answer: `raw` and some `complex` numbers can't appear in an expression, since constructing both requires a function call (`raw()` and `+` respectively). Similarly, creating atomic vectors longer than one requires `c()`.

```{r}
expr(f(TRUE, 1L, 1.0, "x", 1i, raw(1L), 1 + 1i))
expr(c(1, 2))
```

AR Solutions: There is no way to create raws and complex atomics without using a function call (this is only possible for imaginary scalars like `i`, `5i` etc.). But expressions that include a function are *calls*. Therefore, both of these vector types cannot appear in an expression.

Similarly, it is not possible to create an expression that evaluates to an atomic of length greater than one without using a function (e.g. `c()`).

Let's make this observation concrete via an example:

```{r}
# Atomic
is_atomic(expr(1))
# Not an atomic (although it would evaluate to an atomic)
is_atomic(expr(c(1, 1)))
is_call(expr(c(1, 1)))
```

---

2.  What happens when you subset a call object to remove the first element?
    e.g. `expr(read.csv("foo.csv", header = TRUE))[-1]`. Why?

Answer: as the result demonstrates, this removes the function call which makes the first argument the new function call in the expression, since the AST structure (call object) is preserved, and the first element of the call object is the function position.

```{r}
expr(read.csv("foo.csv", header = TRUE))[-1]
```

AR Solutions: When the first element of a call object is removed, the second element moves to the first position, which is the function to call. Therefore, we get `"foo.csv"(header = TRUE)`.

---

3.  Describe the differences between the following call objects.

```{r, results = FALSE}
x <- 1:10
call2(median, x, na.rm = TRUE)
call2(expr(median), x, na.rm = TRUE)
call2(median, expr(x), na.rm = TRUE)
call2(expr(median), expr(x), na.rm = TRUE)
```

Answer: the objects differ in their use of `expr()`. When using bare `median` and `x`, the object uses the evaluated function and value of `x` where using `expr()` prevents evaluation.

```{r}
median
expr(median)
x
expr(x)
```

AR Solutions: The call objects differ in their first two elements, which are in some cases evaluated before the call is constructed. In the first one, both `median()` and `x` are evaluated and inlined into the call.  Therefore, we can see in the constructed call that median is a generic and the `x` argument is 1:10.

```{r}
call2(median, x, na.rm = TRUE)
```

In the following calls we remain with differing combinations. Once, only `x` and once only `median()` gets evaluated. 

```{r}
call2(expr(median), x, na.rm = TRUE)
call2(median, expr(x), na.rm = TRUE)
```

In the final call neither `x` nor `median()` is evaluated.

```{r}
call2(expr(median), expr(x), na.rm = TRUE)
```

Note that all these calls will generate the same result when evaluated. The key difference is when the values bound to the `x` and `median` symbols are found.

---

4.  `rlang::call_standardise()` doesn't work so well for the following calls.
    Why? What makes `mean()` special?

```{r}
call_standardise(quote(mean(1:10, na.rm = TRUE)))
call_standardise(quote(mean(n = T, 1:10))) # nolint: T_and_F_symbol_linter.
call_standardise(quote(mean(x = 1:10, , TRUE))) # nolint: missing_argument_linter.
```

Answer: as noted in the book, "If the function uses `...` it’s not possible to standardise all arguments." `mean()` has a single argument, `x`, and uses `...` for the remaining arguments, although the default S3 method is `mean(x, trim = 0, na.rm = FALSE, ...)`.

AR Solutions: The reason for this unexpected behaviour is that `mean()` uses the `...` argument and therefore cannot standardise the regarding arguments. Since `mean()` uses S3 dispatch (i.e. `UseMethod()`) and the underlying `mean.default()` method specifies some more arguments, `call_standardise()` can do much better with a specific S3 method.

---

5.  Why does this code not make sense?

```{r, eval = FALSE}
x <- expr(foo(x = 1))
names(x) <- c("x", "y")
```

Answer: as the code below demonstrates, the only symbol (name) in the expression is `x`. Assigning a name to both components changes `x` to `y` but doesn't change the function call `foo`.

```{r}
ex <- expr(foo(x = 1))
names(ex)
names(ex) <- c("x", "y")
ex
```

AR Solutions: As stated in *Advanced R*

> The first element of a call is always the function that gets called.

Let's see what happens when we run the code

```{r}
x <- expr(foo(x = 1))
x
names(x) <- c("x", "")
x
names(x) <- c("", "x")
x
```

So, giving the first element a name just adds metadata that R ignores.

---

6.  Construct the expression `if(x > 1) "a" else "b"` using multiple calls to
    `call2()`. How does the code structure reflect the structure of the AST?

Answer: code below. The `call2()` structure mirrors the AST of the if expression.

```{r}
call2("if", call2(">", expr(x), 1), "a", "b")
lobstr::ast(if (x > 1) "a" else "b")
```

AR Solutions: Similar to the prefix version we get

```{r}
call2("if", call2(">", sym("x"), 1), "a", "b")
```

When we read the AST from left to right, we get the same structure: Function to evaluate, expression, which is another function and is evaluated first, and two constants which will be evaluated next.

```{r}
ast(`if`(x > 1, "a", "b"))
```

---

## 18.4.4 Exercises

1.  R uses parentheses in two slightly different ways as illustrated by
    these two calls:

```{r, eval = FALSE}
f((1))
`(`(1 + 1)
```

Compare and contrast the two uses by referencing the AST.

Answer: R uses parentheses both as a function *and* as part of a function call. In the first call, the outer parentheses are for `f` and the inner represent `(` with a single parameter. In the second call, equivalent to `(1 + 1)`, the parentheses are for the `(` function call.

```{r}
lobstr::ast(f((1)))
lobstr::ast(`(`(1 + 1))
lobstr::ast((1 + 1))
```

AR Solutions: The trick with these examples lies in the fact that `(` can be a part of R's general prefix function syntax but can also represent a call to the `(` function.

So, in the AST of the first example, we will not see the outer `(` since it is prefix function syntax and belongs to `f()`. In contrast, the inner `(` is a function (represented as a symbol in the AST):

```{r}
ast(f((1)))
```

In the second example, we can see that the outer `(` is a function and the inner `(` belongs to its syntax:

```{r}
ast(`(`(1 + 1))
```

For the sake of clarity, let's also create a third example, where none of the `(` is part of another function's syntax:

```{r}
ast(((1 + 1)))
```

---

2.  `=` can also be used in two ways. Construct a simple example that shows
    both uses.

Answer: `=` can be used for assignment and also for function parameters.

```{r}
lobstr::ast(`=`(val, sum(10, na.rm = TRUE)))
```

AR Solutions: `=` is used both for assignment, and for naming arguments in function calls:

```{r}
b = c(c = 1) # nolint: assignment_linter. styler: off
```

So, when we play with `ast()`, we can directly see that the following is not possible:

```{r, error = TRUE}
try(ast(b = c(c = 1)))
```

We get an error because `b = ` makes R looking for an argument called `b`. Since `x` is the only argument of `ast()`, we get an error.

The easiest way around this problem is to wrap this line in `{}`.

```{r}
ast({b = c(c = 1)})  # nolint: assignment_linter, brace_linter. styler: off
```

When we ignore the braces and compare the trees, we can see that the first `=` is used for assignment and the second `=` is part of the syntax of function calls.

---

3.  Does `-2^2` yield 4 or -4? Why?

Answer: the result is -4, because `^` has higher precedence than `-`:

```{r}
-2^2
lobstr::ast(-2^2)
```

AR Solutions: It yields `-4`, because `^` has a higher operator precedence than `-`, which we can verify by looking at the AST (or looking it up under `?"Syntax"`):

```{r}
-2^2

ast(-2^2)
```

---

4.  What does `!1 + !1` return? Why?

Answer: `FALSE`. As the AST demonstrates, the `!` operator takes precedence.

```{r}
!1 + !1
lobstr::ast(!1 + !1)
```

The result can be demonstrated by evaluating the AST in steps:

```{r}
!1 # FALSE
1 + FALSE # 1
!1 # FALSE
```

AR Solutions: The answer is a little surprising:

```{r}
!1 + !1
```

To answer the "why?", we take a look at the AST:

```{r}
ast(!1 + !1)
```

The right `!1` is evaluated first. It evaluates to `FALSE`, because R coerces every non 0 numeric to `TRUE`, when a logical operator is applied. The negation of `TRUE` then equals `FALSE`.

Next `1 + FALSE` is evaluated to `1`, since `FALSE` is coerced to `0`.

Finally `!1` is evaluated to `FALSE`.

Note that if `!` had a higher precedence, the intermediate result would be `FALSE + FALSE`, which would evaluate to `0`.

---

5.  Why does `x1 <- x2 <- x3 <- 0` work? Describe the two reasons.

Answer: assignment is right-associative: `x1 <- (x2 <- 0)` and `x1 <- x2 <- 0` are equivalent, and the rightmost assignment operator is evaluated first.

```{r}
lobstr::ast(x1 <- x2 <- x3 <- 0)
```

AR Solutions: One reason is that `<-` is right-associative, i.e. evaluation takes place from right to left:

```{r}
x1 <- (x2 <- (x3 <- 0))
```

The other reason is that `<-` invisibly returns the value on the right-hand side.

```{r}
(x3 <- 0)
```

Note: I forgot that `<-` returns the value invisibly.

---

6.  Compare the ASTs of `x + y %+% z` and `x ^ y %+% z`. What have you learned
    about the precedence of custom infix functions?

Answer: custom infix functions have higher precedence than `+` and lower precedence than `^`.

```{r}
lobstr::ast(x + y %+% z)
lobstr::ast(x ^ y %+% z) # styler: off
```

AR Solutions: Let's take a look at the syntax trees:

```{r}
ast(x + y %+% z)
```

Here `y %+% z` will be calculated first and the result will be added to `x`.

```{r}
ast(x ^ y %+% z) # styler: off
```

Here `x ^ y` will be calculated first, and the result will be used as first argument to `%+%()`.
We can conclude that custom infix functions have precedence between addition and exponentiation.
The exact precedence of infix functions can be looked up under `?"Syntax"` where we see that it lies directly behind the sequence operator (`:`) and in front of the multiplication and division operators (`*` and `/`).

---

7.  What happens if you call `parse_expr()` with a string that generates
    multiple expressions? e.g. `parse_expr("x + 1; y + 1")`

Answer: `parse_expr()` throws an error. Use `parse_exprs()` with strings containing multiple expressions.

```{r}
try(parse_expr("x + 1; y + 1"))
parse_exprs("x + 1; y + 1")
```

AR Solutions: In this case `parse_expr()` notices that more than one expression would have to be generated and throws an error.

```{r, error = TRUE}
try(parse_expr("x + 1; y + 1"))
```

---

8.  What happens if you attempt to parse an invalid expression? e.g. `"a +"`
    or `"f())"`.

Answer: `parse_expr()` throws an error.

```{r}
try(parse_expr("a +"))
try(parse_expr("f())"))
```

AR Solutions: Invalid expressions will lead to an error in the underlying `parse()` function.

```{r, error = TRUE}
try(parse_expr("a +"))
try(parse_expr("f())"))
try(parse(text = "a +"))
try(parse(text = "f())"))
```

---

9.  `deparse()` produces vectors when the input is long. For example, the
    following call produces a vector of length two:

```{r}
expr <- expr(g(a + b + c + d + e + f + g + h + i + j + k + l +
  m + n + o + p + q + r + s + t + u + v + w + x + y + z))
deparse(expr)
```

What does `expr_text()` do instead?

Answer: `expr_text()` adds a newline (`\n`) instead.

```{r}
expr_text(expr)
```

AR Solutions: `expr_text()` will paste the results from `deparse(expr)` together and use a linebreak (`\n`) as separator.

```{r, eval = FALSE}
expr <- expr(g(a + b + c + d + e + f + g + h + i + j + k + l + m +
  n + o + p + q + r + s + t + u + v + w + x + y + z))
deparse(expr)
#> [1] "g(a + b + c + d + e + f + g + h + i + j + k + l + m + n + "
#> [2] "o + p + q + r + s + t + u + v + w + x + y + z)"
expr_text(expr)
#> [1] "g(a + b + c + d + e + f + g + h + i + j + k + l + m + n
#> + \n    o + p + q + r + s + t + u + v + w + x + y + z)"
```

---

10.  `pairwise.t.test()` assumes that `deparse()` always returns a length one
    character vector. Can you construct an input that violates this expectation?
    What happens?

Answer: an especially long input for `x` or `g` would create a character vector of length two or more. However, starting in R 4.0.0, `pairwise.t.test()` uses `deparse1()` which guarantees a string (character of length one), so it is no longer possible to break `pairwise.t.test()`.

```{r}
pairwise.t.test
deparse(c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18))
pairwise.t.test(1:18, c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18))
```

AR Solutions: The function `pairwise.t.test()` captures its data arguments (`x` and `g`) so it can print the input expressions along the computed p-values. Prior to R 4.0.0 this used to be implemented via `deparse(substitute(x))` in combination with `paste()`. This could lead to unexpected output, if one of the inputs exceeded the default `width.cutoff` value of 60 characters within `deparse()`. In this case, the expression would be split into a character vector of length greater 1.

```{r, eval = FALSE}
# Output in R version 3.6.2
d <- 1
pairwise.t.test(2, d + d + d + d + d + d + d + d +
  d + d + d + d + d + d + d + d + d)
#> 	Pairwise comparisons using t tests with pooled SD
#>
#> data:  2 and d + d + d + d + d + d + d + d + d + d + d + d + d + d
#> + d + d +  2 and     d
#>
#> <0 x 0 matrix>
#>
#> P value adjustment method: holm
```

In [R 4.0.0](https://cran.r-project.org/doc/manuals/r-release/NEWS.html) `pairwise.t.test()` was updated to use the newly introduced `deparse1()`, which serves as a wrapper around `deparse()`.

> deparse1() is a simple utility added in R 4.0.0 to ensure a string result (character vector of length one), typically used in name construction, as deparse1(substitute(.)).

```{r, eval = FALSE}
# Output since R 4.0.0
d <- 1
pairwise.t.test(2, d + d + d + d + d + d + d + d +
  d + d + d + d + d + d + d + d + d)
#> 	Pairwise comparisons using t tests with pooled SD
#>
#> data:  2 and d + d + d + d + d + d + d + d + d + d + d + d + d + d
#> + d + d + d
#>
#> <0 x 0 matrix>
#>
#> P value adjustment method: holm
```

---

```{r}
expr_type <- function(x) {
  if (rlang::is_syntactic_literal(x)) {
    "constant"
  } else if (is.symbol(x)) {
    "symbol"
  } else if (is.call(x)) {
    "call"
  } else if (is.pairlist(x)) {
    "pairlist"
  } else {
    typeof(x)
  }
}

switch_expr <- function(x, ...) {
  switch(expr_type(x),
    ...,
    stop("Don't know how to handle type ", typeof(x), call. = FALSE)
  )
}
```

## 18.5.3 Exercises

1.  `logical_abbr()` returns `TRUE` for `T(1, 2, 3)`. How could you modify
    `logical_abbr_rec()` so that it ignores function calls that use `T` or `F`?

AR Solutions: We can apply a similar logic as in the [assignment example](https://adv-r.hadley.nz/expressions.html#finding-all-variables-created-by-assignment) from *Advanced R*. We just treat it as a special case handled within a sub function called `find_T_call()`, which finds `T()` calls and "bounces them out". Therefore, we also repeat the `expr_type()` helper which tells us if we are in the base or in the recursive case.

```{r}
expr_type <- function(x) {
  if (rlang::is_syntactic_literal(x)) {
    "constant"
  } else if (is.symbol(x)) {
    "symbol"
  } else if (is.call(x)) {
    "call"
  } else if (is.pairlist(x)) {
    "pairlist"
  } else {
    typeof(x)
  }
}
switch_expr <- function(x, ...) {
  switch(expr_type(x),
    ...,
    stop("Don't know how to handle type ",
      typeof(x),
      call. = FALSE
    )
  )
}
```

```{r}
find_T_call <- function(x) { # nolint: object_name_linter.
  if (is_call(x, "T")) {
    x <- as.list(x)[-1]
    purrr::some(x, logical_abbr_rec)
  } else {
    purrr::some(x, logical_abbr_rec)
  }
}
logical_abbr_rec <- function(x) {
  switch_expr(
    x,
    # Base cases
    constant = FALSE,
    symbol = as_string(x) %in% c("F", "T"),

    # Recursive cases
    pairlist = purrr::some(x, logical_abbr_rec),
    call = find_T_call(x)
  )
}
logical_abbr <- function(x) {
  logical_abbr_rec(enexpr(x))
}
```

Now let's test our new `logical_abbr()` function:

```{r}
logical_abbr(T(1, 2, 3))
logical_abbr(T(T, T(3, 4))) # nolint: T_and_F_symbol_linter.
logical_abbr(T(T)) # nolint: T_and_F_symbol_linter.
logical_abbr(T())
logical_abbr()
logical_abbr(c(T, T, T)) # nolint: T_and_F_symbol_linter.
```

Answer: taken from Advanced R solutions, we need to find function calls within the recursive call. The trick is to find and remove the `T` or `F` from the list:

```{r}
find_tf_call <- function(x) {
  if (is_call(x, "T") || is_call(x, "F")) {
    x <- as.list(x)[-1]
    purrr::some(x, logical_abbr_rec)
  } else {
    purrr::some(x, logical_abbr_rec)
  }
}

logical_abbr_rec <- function(x) {
  switch_expr(
    x,
    # Base cases
    constant = FALSE,
    symbol = as_string(x) %in% c("F", "T"),

    # Recursive cases
    pairlist = purrr::some(x, logical_abbr_rec),
    call = find_tf_call(x)
  )
}

logical_abbr <- function(x) {
  logical_abbr_rec(enexpr(x))
}

logical_abbr(T) # TRUE  # nolint: T_and_F_symbol_linter.
logical_abbr(FALSE) # FALSE
logical_abbr(T(1, 2, 3)) # FALSE
logical_abbr(F(1, 2, 3)) # FALSE
```

Note: I was unable to solve this exercise on my own, and needed to consult AR Solutions. The AR Solutions version only implements `T`, adding `F` is trivial.

---

2.  `logical_abbr()` works with expressions. It currently fails when you give it
    a function. Why? How could you modify `logical_abbr()` to make it
    work? What components of a function will you need to recurse over?

```{r, eval = FALSE}
logical_abbr(function(x = TRUE) {
  g(x + T) # nolint: T_and_F_symbol_linter.
})
```

AR Solutions: The function currently fails, because `"closure"` is not handled in `switch_expr()` within `logical_abbr_rec()`.

```{r, error = TRUE}
f <- function(x = TRUE) {
  g(x + T) # nolint: object_usage_linter, T_and_F_symbol_linter.
}
try(logical_abbr(!!f))
```

If we want to make it work, we have to write a function to also iterate over the formals and the body of the input function.

Answer: the above example evaluates successfully as `TRUE`, however, the following fails, presumably because `expr_type` doesn't handle `closure`:

```{r}
try(logical_abbr(function(x = TRUE) {
  g(x)
}))

is_closure(function(x = TRUE) {
  g(x)
})

typeof(function(x = TRUE) {
  g(x)
})
```

Note: I'm not sure why the error says `integer` instead of `closure`.

---

3.  Modify `find_assign` to also detect assignment using replacement
    functions, i.e. `names(x) <- y`.

Answer: first compare the AST to normal assignment:

```{r}
ast(x <- y)
ast(names(x) <- y)
```

Assignment is a call object where the first element is the symbol `<-`, the second is a call object with the function and the name of the variable, and the third is the value to be assigned. The original code doesn't detect `name(x) <- y`:

```{r}
flat_map_chr <- function(.x, .f, ...) {
  purrr::flatten_chr(purrr::map(.x, .f, ...))
}

find_assign_call <- function(x) {
  if (is_call(x, "<-") && is_symbol(x[[2]])) {
    lhs <- as_string(x[[2]])
    children <- as.list(x)[-1]
  } else {
    lhs <- character()
    children <- as.list(x)
  }

  c(lhs, flat_map_chr(children, find_assign_rec))
}

find_assign_rec <- function(x) {
  switch_expr(x,
    # Base cases
    constant = , # nolint: missing_argument_linter.
    symbol = character(),

    # Recursive cases
    pairlist = flat_map_chr(x, find_assign_rec),
    call = find_assign_call(x)
  )
}

find_assign <- function(x) unique(find_assign_rec(enexpr(x)))

find_assign(a <- b <- c <- 1)
find_assign(names(x) <- y)
```

To detect the new pattern, modify `find_assign_call` to detect when the second element is a call, and return the third:

```{r}
find_assign_call <- function(x) {
  if (is_call(x, "<-") && is_symbol(x[[2]])) {
    lhs <- as_string(x[[2]])
    children <- as.list(x)[-1]
  } else if (is_call(x, "<-") && is_call(x[[2]])) {
    lhs <- as_string(x[[3]])
    children <- as.list(x)[-1]
  } else {
    lhs <- character()
    children <- as.list(x)
  }

  c(lhs, flat_map_chr(children, find_assign_rec))
}

find_assign(a <- b <- c <- 1)
find_assign(names(x) <- y)
```

AR Solutions: Let's see what the AST of such an assignment looks like:

```{r}
ast(names(x) <- x)
```

So, we need to catch the case where the first two elements are both calls. Further the first call is identical to `<-` and we must return only the second call to see which objects got new values assigned.

This is why we add the following block within another `else` statement in `find_assign_call()`:

```{r, eval = FALSE}
if (is_call(x, "<-") && is_call(x[[2]])) {
  lhs <- expr_text(x[[2]])
  children <- as.list(x)[-1]
}
```

Let us finish with the whole code, followed by some tests for our new function:

```{r}
flat_map_chr <- function(.x, .f, ...) {
  purrr::flatten_chr(purrr::map(.x, .f, ...))
}

find_assign <- function(x) unique(find_assign_rec(enexpr(x)))

find_assign_call <- function(x) {
  if (is_call(x, "<-") && is_symbol(x[[2]])) {
    lhs <- as_string(x[[2]])
    children <- as.list(x)[-1]
  } else {
    if (is_call(x, "<-") && is_call(x[[2]])) {
      lhs <- expr_text(x[[2]])
      children <- as.list(x)[-1]
    } else {
      lhs <- character()
      children <- as.list(x)
    }
  }

  c(lhs, flat_map_chr(children, find_assign_rec))
}

find_assign_rec <- function(x) {
  switch_expr(
    x,
    # Base cases
    constant = , symbol = character(), # nolint: missing_argument_linter.
    # Recursive cases
    pairlist = flat_map_chr(x, find_assign_rec),
    call = find_assign_call(x)
  )
}

# Tests functionality
find_assign(x <- y)
find_assign(names(x))
find_assign(names(x) <- y)
find_assign(names(x(y)) <- y)
find_assign(names(x(y)) <- y <- z)
```

Note: using `expr_text()` instead of `as_string()` is preferable.

---

4.  Write a function that extracts all calls to a specified function.

Answer: this is similar to the previous exercise. Here is a simple version that doesn't properly walk the tree:

```{r}
find_fun_call <- function(x, fun) {
  if (is_call(x, fun)) expr_text(x) else character()
}

find_call_rec <- function(x, fun) {
  switch_expr(
    x,
    # Base cases
    constant = , # nolint: missing_argument_linter.
    symbol = character(),

    # Recursive cases
    pairlist = flat_map_chr(x, find_call_rec),
    call = find_fun_call(x, fun)
  )
}

find_call_fun <- function(x, fun) find_call_rec(enexpr(x), fun)

find_call_fun(1:10, "sum")
find_call_fun(x, "sum")
find_call_fun(sum(1:10), "sum")
find_call_fun(sum(1:10), "mean")
find_call_fun(sum(1:10) + sum(11:20), "sum")
```

Updated version:

```{r}
find_fun_call <- function(x, fun) {
  if (is_call(x, fun)) {
    parents <- expr_text(x)
    children <- as.list(x)[-1]
  } else {
    parents <- character()
    children <- as.list(x)
  }

  c(parents, flat_map_chr(children, find_call_rec, fun))
}

find_call_fun(1:10, "sum")
find_call_fun(x, "sum")
find_call_fun(sum(1:10), "sum")
find_call_fun(sum(1:10), "mean")
find_call_fun(sum(1:10) + sum(11:20), "sum")
find_call_fun(sum(sum(1:10), sum(11:20)), "sum")
```

AR Solutions: Here we need to delete the previously added else statement and check for a call (not necessarily `<-`) within the first `if()` in `find_assign_call()`. We save a call when we found one and return it later as part of our character output. Everything else stays the same:

```{r}
find_assign_call <- function(x) {
  if (is_call(x)) {
    lhs <- expr_text(x)
    children <- as.list(x)[-1]
  } else {
    lhs <- character()
    children <- as.list(x)
  }

  c(lhs, flat_map_chr(children, find_assign_rec))
}

find_assign_rec <- function(x) {
  switch_expr(
    x,
    # Base cases
    constant = , # nolint: missing_argument_linter.
    symbol = character(),

    # Recursive cases
    pairlist = flat_map_chr(x, find_assign_rec),
    call = find_assign_call(x)
  )
}

find_assign(x <- y)
find_assign(names(x(y)) <- y <- z)
find_assign(mean(sum(1:3)))
```

Note: the AR Solutions version detects *any* function call, not a specific function call.

---

# 19 Quasiquotation

Now that you understand the tree structure of R code, it's time to return to one of the fundamental ideas that make `expr()` and `ast()` work: quotation. In tidy evaluation, all quoting functions are actually quasiquoting functions because they also support unquoting. Where quotation is the act of capturing an unevaluated expression, __unquotation__ is the ability to selectively evaluate parts of an otherwise quoted expression. Together, this is called quasiquotation. Quasiquotation makes it easy to create functions that combine code written by the function's author with code written by the function's user. This helps to solve a wide variety of challenging problems. 

Quasiquotation is one of the three pillars of tidy evaluation. You'll learn about the other two (quosures and the data mask) in Chapter 20. When used alone, quasiquotation is most useful for programming, particularly for generating code. But when it's combined with the other techniques, tidy evaluation becomes a powerful tool for data analysis.

## 19.2.2 Exercises

1.  For each function in the following base R code, identify which arguments
    are quoted and which are evaluated.

```{r, results = FALSE}
library(MASS) # nolint: unused_import_linter.

mtcars2 <- subset(mtcars, cyl == 4)

with(mtcars2, sum(vs))
sum(mtcars2$am)

rm(mtcars2)
```

Answer: using the technique described in the chapter, test arguments outside their functions:

```{r}
library(MASS) # nolint: unused_import_linter.
try(MASS)

mtcars2 <- subset(mtcars, cyl == 4)
mtcars
try(cyl)

with(mtcars2, sum(vs))
mtcars2
try(vs)

sum(mtcars2$am)
mtcars2$am

rm(mtcars2)
```

Results: `MASS`, `cyl`, `vs` are quoted, `mtcars`, `mtcars2` and `mtcars2$am` are not.

AR Solutions: For each argument we first follow the advice from *Advanced R* and execute the argument outside of the respective function. Since `MASS`, `cyl`, `vs` and `am` are not objects contained in the global environment, their execution raises an "Object not found" error. This way we confirm that the respective function arguments are quoted. For the other arguments, we may inspect the source code (and the documentation) to check if any quoting mechanisms are applied or the arguments are evaluated.

```{r, eval = FALSE}
library(MASS) # MASS is quoted # nolint: unused_import_linter.
```

`library()` also accepts character vectors and doesn't quote when `character.only` is set to `TRUE`, so `library(MASS, character.only = TRUE)` would raise an error.

```{r, eval = FALSE}
mtcars2 <- subset(mtcars, cyl == 4) # mtcars is evaluated
# cyl is quoted
with(mtcars2, sum(vs)) # mtcars2 is evaluated
# sum(vs) is quoted
sum(mtcars2$am) # matcars$am is evaluated
# am is quoted by $()
```

When we inspect the source code of `rm()`, we notice that `rm()` catches its `...` argument as an unevaluated call (in this case a pairlist) via `match.call()`. This call is then converted into a string for further evaluation.

```{r, eval = FALSE}
rm(mtcars2) # mtcars2 is quoted
```

---

2.  For each function in the following tidyverse code, identify which arguments
    are quoted and which are evaluated.

```{r, eval = FALSE}
library(dplyr)
library(ggplot2)

by_cyl <- mtcars %>%
  group_by(cyl) %>%
  summarise(mean = mean(mpg))

ggplot(by_cyl, aes(cyl, mean)) +
  geom_point()
```

Answer: `library` automatically quotes its arguments. `cyl`, `mpg`, and `mean` are all data.frame columns.

- Quoted: `dplyr`, `ggplot2`, `cyl`, `mpg`, `mean` (in the final `ggpot` call)
- Evaluated: `by_cyl`, `mtcars`

AR Solutions: From the previous exercise we've already learned that `library()` quotes its first argument.

```{r, eval = FALSE}
library(dplyr) # dplyr is quoted
library(ggplot2) # ggplot2 is quoted
```

In similar fashion, it becomes clear that `cyl` is quoted by `group_by()`.

```{r, eval = FALSE}
by_cyl <- mtcars %>% # mtcars is evaluated
  group_by(cyl) %>% # cyl is quoted
  summarise(mean = mean(mpg)) # mean = mean(mpg) is quoted
```

To find out what happens in `summarise()`, we inspect the source code. Tracing down the S3-dispatch of `summarise()`, we see that the `...` argument is quoted in `dplyr:::summarise_cols()` which is called in the underlying `summarise.data.frame()` method.

```{r}
dplyr::summarise
```

```{r}
dplyr:::summarise.data.frame # nolint: undesirable_operator_linter.
```

```{r, eval = FALSE}
dplyr:::summarise_cols # nolint: undesirable_operator_linter.
#> function (.data, ...)
#> {
#>     mask <- DataMask$new(.data, caller_env())
#>     dots <- enquos(...)
#>     dots_names <- names(dots)
#>     auto_named_dots <- names(enquos(..., .named = TRUE))
#>     cols <- list()
#>     sizes <- 1L
#>     chunks <- vector("list", length(dots))
#>     types <- vector("list", length(dots))
#>
#>     ## function definition abbreviated for clarity ##
#> }
#> <bytecode: 0x55b540c07ca0>
#> <environment: namespace:dplyr>
```

In the following `{ggplot2}` expression the `cyl`- and `mean`-objects are quoted.

```{r, eval = FALSE}
ggplot(
  by_cyl, # by_cyl is evaluated
  aes(cyl, mean) # aes() is evaluated
) +
  # cyl, mean is quoted (via aes)
  geom_point()
```

We can confirm this also by inspecting `aes()`'s source code.

```{r}
ggplot2::aes
```

---

## 19.3.6 Exercises

1.  How is `expr()` implemented? Look at its source code.

Answer: reviewing the source code:

```{r}
rlang::expr
rlang::enexpr
```

`expr()` simply calls `enexpr()`, which uses the C++ function `ffi_enexpr` and uses the base R function `substitute()`.

AR Solutions: `expr()` acts as a simple wrapper, which passes its argument to `enexpr()`.

---

2.  Compare and contrast the following two functions. Can you predict the
    output before running them?

```{r, results = FALSE}
f1 <- function(x, y) {
  exprs(x = x, y = y)
}
f2 <- function(x, y) {
  enexprs(x = x, y = y)
}
f1(a + b, c + d)
f2(a + b, c + d)
```

Answer: per Advanced R, `exprs(x = x, y = y)` is shorthand for `list(x = expr(x), y = expr(y))`, and `expr()` will capture the argument exactly as provided. `enexpr()` captures what the caller supplied, with `enexprs()` returning a list.

Test the predictions in comments:

```{r}
f1 <- function(x, y) {
  exprs(x = x, y = y)
}
f2 <- function(x, y) {
  enexprs(x = x, y = y)
}
# returns $x x $y y
f1(a + b, c + d)
# returns $x a + b $y c + d
f2(a + b, c + d)
```

Correct!

AR Solutions: Both functions are able to capture multiple arguments and will return a named list of expressions. `f1()` will return the arguments defined within the body of `f1()`. This happens because `exprs()` captures the expressions as specified by the developer during the definition of `f1()`.

```{r}
f1(a + b, c + d)
```

`f2()` will return the arguments supplied to `f2()` as specified by the user when the function is called.

```{r}
f2(a + b, c + d)
```

---

3.  What happens if you try to use `enexpr()` with an expression (i.e. 
    `enexpr(x + y)` ? What happens if `enexpr()` is passed a missing argument?

Answer: from the code below, `arg` must be a symbol, not an expression, and `arg` must exist.

```{r}
try(enexpr(x + y))
try(enexpr(arg = foo))
```

AR Solutions: In the first case an error is thrown:

```{r, error = TRUE}
on_expr <- function(x) {
  enexpr(expr(x))
}
try(on_expr(x + y))
```

In the second case a missing argument is returned:

```{r}
on_missing <- function(x) {
  enexpr(x)
}
on_missing()
is_missing(on_missing())
```

---

4.  How are `exprs(a)` and `exprs(a = )` different? Think about both the
    input and the output.

Answer: as the code below demonstrates, `expr(a)` creates an unnamed list with `a` as the first element. `exprs(a = )` which creates an named list with an empty first element. Input of the form `x = y` creates a named list, while the form `x` creates an unnamed list.

```{r}
exprs(a)
exprs(a = ) # nolint: missing_argument_linter, spaces_inside_linter.

exprs(a = a)
```

AR Solutions: In `exprs(a)` the input `a` is interpreted as a symbol for an unnamed argument. Consequently, the output shows an unnamed list with the first element containing the symbol `a`.

```{r}
out1 <- exprs(a)
str(out1)
```

In `exprs(a = )` the first argument is named `a`, but then no value is provided. This leads to the output of a named list with the first element named `a`, which contains the missing argument.

```{r}
out2 <- exprs(a = ) # nolint: missing_argument_linter, spaces_inside_linter.
str(out2)
is_missing(out2$a)
```

---

5.  What are other differences between `exprs()` and `alist()`? Read the 
    documentation for the named arguments of `exprs()` to find out.

Answer: `exprs()` has three named arguments (`alist()` has none):

- `.named`, which automatically adds names to the list
- `.ignore_empty`, which selectively ignores empty arguments depending on the setting
- `.unquote_names`, which treats use of `:=` (which allows names injection) as `=`

AR Solutions: `exprs()` provides the additional arguments `.named` (`= FALSE`), `.ignore_empty` (`c("trailing", "none", "all")`) and `.unquote_names` (`TRUE`). `.named` allows to ensure that all dots are named. `ignore_empty` allows to specify how empty arguments should be handled for dots (`"trailing"`) or all arguments (`"none"` and `"all"`). Further via `.unquote_names` one can specify if `:=` should be treated like `=`. `:=` can be useful as it supports unquoting (`!!`) on the left-hand side.

---

6.  The documentation for `substitute()` says:

    > Substitution takes place by examining each component of the parse tree 
    > as follows: 
    > 
    > * If it is not a bound symbol in `env`, it is unchanged. 
    > * If it is a promise object (i.e., a formal argument to a function) 
    >   the expression slot of the promise replaces the symbol. 
    > * If it is an ordinary variable, its value is substituted, unless 
    > `env` is `.GlobalEnv` in which case the symbol is left unchanged.
  
    Create examples that illustrate each of the above cases.

Answer: reviewing Google results for [R bound symbol](https://rlang.r-lib.org/reference/env_names.html), [R promise object](https://rstudio.github.io/promises/articles/overview.html), and chapter 6.5.1, construct the following code:

```{r}
# ordinary variable in `.GlobalEnv`
foo <- "bar"
substitute(foo)
# ordinary variable not in `.GlobalEnv`
local({
  foo <- "bar"
  substitute(foo)
})
# not a bound symbol in `env`
substitute(func)
# promise object
local({
  x <- 1:5
  y <- 1:10
  substitute(mean(x = y))
})
```

In the promise example, `y` is substituted, but not `x`, since `x` is the formal argument to `mean()`.

AR Solutions: Let's create a new environment `my_env`, which contains no objects. In this case `substitute()` will just return its first argument (`expr`):

```{r}
my_env <- env()
substitute(x, my_env)
```

When we create a function containing an argument, which is directly returned after substitution, this function just returns the provided expression:

```{r}
foo <- function(x) substitute(x)

foo(x + y * sin(0))
```

In case `substitute()` can find (parts of) the expression in `env`, it will literally substitute. However, unless `env` is `.GlobalEnv`.

```{r}
my_env$x <- 7
substitute(x, my_env)

x <- 7
substitute(x, .GlobalEnv)
```

---

## 19.4.8 Exercises

1.  Given the following components:

```{r}
xy <- expr(x + y)
xz <- expr(x + z)
yz <- expr(y + z)
abc <- exprs(a, b, c)
```

Use quasiquotation to construct the following calls:

```{r, eval = FALSE}
(x + y) / (y + z)
-(x + z)^(y + z)
(x + y) + (y + z) - (x + y)
atan2(x + y, y + z)
sum(x + y, x + y, y + z)
sum(a, b, c)
mean(c(a, b, c), na.rm = TRUE)
foo(a = x + y, b = y + z)
```

Answer: code below.

```{r}
# nolint start: commented_code_linter.
# (x + y) / (y + z)
expr(!!xy / !!yz)

# -(x + z) ^ (y + z)
expr(-(!!xz)^!!yz)

# (x + y) + (y + z) - (x + y)
expr(((!!xy)) + !!yz - !!xy)

# atan2(x + y, y + z)
expr(atan2(!!xy, !!yz))

# sum(x + y, x + y, y + z)
expr(sum(!!xy, !!xy, !!yz))

# sum(a, b, c)
expr(sum(!!!abc))

# mean(c(a, b, c), na.rm = TRUE)
expr(mean(c(!!!abc), na.rm = TRUE)) # nolint: unneeded_concatenation_linter.

# foo(a = x + y, b = y + z)
expr(foo(a = !!xy, b = !!yz))

# nolint end
```

Surprisingly, multiple parentheses are needed to construct `(x + y) + (y + z) - (x + y)`.

AR Solutions We combine and unquote the given quoted expressions to construct the desired calls like this:

```{r}
expr(!!xy / !!yz) # (1)

expr(-(!!xz)^(!!yz)) # (2)

expr(((!!xy)) + !!yz - !!xy) # (3)

expr(atan2(!!xy, !!yz)) # (4)

expr(sum(!!xy, !!xy, !!yz)) # (5)

expr(sum(!!!abc)) # (6)

expr(mean(c(!!!abc), na.rm = TRUE)) # (7)  # nolint: unneeded_concatenation_linter.

expr(foo(a = !!xy, b = !!yz)) # (8)
```

---

2.  The following two calls print the same, but are actually different:

```{r}
(a <- expr(mean(1:10)))
(b <- expr(mean(!!(1:10))))
identical(a, b)
```

What's the difference? Which one is more natural?

Answer: the key difference is the use of `!!` for the sequence `1:10`. Comparing `expr()` for both of the arguments supplied to mean, `!!` changes the expression from a call to `:` to an integer vector:

```{r}
arga <- expr(1:10)
argb <- expr(!!(1:10))
arga
ast(!!arga)
argb
ast(!!argb)
```

Use of `!!` creates a non-standard AST with an inline integer. The first form is more natural.

AR Solutions: It's easiest to see the difference with `lobstr::ast()`:

```{r}
lobstr::ast(mean(1:10))
lobstr::ast(mean(!!(1:10)))
```

In the expression `mean(!!(1:10))` the call `1:10` is evaluated to an integer vector, while still being a call object in `mean(1:10)`.

The first version (`mean(1:10)`) seems more natural. It captures lazy evaluation, with a promise that is evaluated when the function is called. The second version (`mean(!!(1:10))`) inlines a vector directly into a call.

---

## 19.6.5 Exercises

1.  One way to implement `exec()` is shown below. Describe how it works. What are the
    key ideas?

```{r}
exec <- function(f, ..., .env = caller_env()) {
  args <- list2(...)
  do.call(f, args, envir = .env)
}
```

Answer: the implementation is a wrapper to `do.call()`, which requires a function, list of arguments, and environment (it also has a `quote` parameter which is set to `FALSE` by default). `list2()` is used to convert the function arguments into a list which is compatible with `do.call()`.

AR Solutions: `exec()` takes a function (`f`), its arguments (`...`) and an environment (`.env`) as input. This allows to construct a call from `f` and `...` and evaluate this call in the supplied environment. As the `...` argument is handled via `list2()`, `exec()` supports tidy dots (quasiquotation), which means that arguments and names (on the left-hand side of `:=`) can be unquoted via `!!` and `!!!`.

---

2.  Carefully read the source code for `interaction()`, `expand.grid()`, and 
    `par()`.  Compare and contrast the techniques they use for switching 
    between dots and list behaviour.

Answer: reviewing the source code for each function, all three functions convert dots into a list using `list(...)`. All three also allow passing multiple arguments as well as a list as a single argument.

```{r}
interaction
expand.grid
par # nolint: undesirable_function_linter.
```

AR Solutions:  All three functions capture the dots via `args <- list(...)`.

`interaction()` computes factor interactions between the captured input factors by iterating over the `args`. When a list is provided this is detected via `length(args) == 1 && is.list(args[[1]])` and one level of the list is stripped through `args <- args[[1]]`. The rest of the function's code doesn't differentiate further between list and dots behaviour.

```{r}
# styler: off
# Both calls create the same output
interaction(     a = c("a", "b", "c", "d"), b = c("e", "f"))  # dots # nolint: spaces_inside_linter.
interaction(list(a = c("a", "b", "c", "d"), b = c("e", "f"))) # list
# styler: on
```

`expand.grid()` uses the same strategy and also assigns `args <- args[[1]]` in case of `length(args) == 1 && is.list(args[[1]])`.

`par()` does the most pre-processing to ensure a valid structure of the `args` argument. When no dots are provided (`!length(args)`) it creates a list of arguments from an internal character vector (partly depending on its `no.readonly` argument). Further, given that all elements of `args` are character vectors (`all(unlist(lapply(args, is.character)))`), `args` is turned into a list via `as.list(unlist(args))` (this flattens nested lists). Similar to the other functions, one level of `args` gets stripped via `args <- args[[1L]]`, when `args` is of length one and its first element is a list.

---

3.  Explain the problem with this definition of `set_attr()`

```{r, error = TRUE}
set_attr <- function(x, ...) {
  attr <- rlang::list2(...)
  attributes(x) <- attr
  x
}
try(set_attr(1:10, x = 10))
```

Answer: `attributes(x) <- value` requires `value` to be a named list, which is only generated for `list2` calls of the form `list2(a = 1, b = 2)` - setting `.named = TRUE` doesn't help in this case as it generates zero-length names which are not allowed by `attributes()`. Using dots will only work when each attribute is explicitly named.

```{r}
set_attr <- function(x, ...) {
  attr <- rlang::list2(..., .named = TRUE)
  attributes(x) <- attr
  x
}
try(set_attr(1:10, x = 10))
```

AR Solutions: `set_attr()` expects an object named `x` and its attributes, supplied via the dots. Unfortunately, this prohibits us to provide attributes named `x` as these would collide with the argument name of our object. Even omitting the object's argument name doesn't help in this case — as can be seen in the example where the object is consequently treated as an unnamed attribute.

However, we may name the first argument `.x`, which seems clearer and less likely to invoke errors. In this case `1:10` will get the (named) attribute `x = 10` assigned:

```{r}
set_attr <- function(.x, ...) {
  attr <- rlang::list2(...)

  attributes(.x) <- attr
  .x
}

set_attr(1:10, x = 10)
```

---
