---
title: "Advanced R (Object-oriented programming)"
author: "John Benninghoff"
date: '2022-08-15'
file-modified: '2023-01-14'
categories: [exercises, advanced-r]
order: 203
output:
  html_notebook:
    theme:
      version: 5
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
---

Workbook for completing quizzes and exercises from the "Object-oriented programming" chapters of [Advanced R](https://adv-r.hadley.nz/index.html), second edition, with comparisons to solutions from [Advanced R Solutions](https://advanced-r-solutions.rbind.io).

```{r setup, message = FALSE, warning = FALSE}
library(sloop)
library(purrr)
library(dplyr)
library(R6)
library(methods)

# from https://github.com/hadley/adv-r/blob/master/common.R
knitr::opts_chunk$set(
  comment = "#>",
  fig.align = "center"
)

knitr::knit_hooks$set(
  small_mar = function(before, options, envir) {
    if (before) {
      par(mar = c(4.1, 4.1, 0.5, 0.5))
    }
  }
)
```

# Introduction

This workbook includes answers and solutions to the quizzes and exercises from [Advanced R](https://adv-r.hadley.nz/index.html) and [Advanced R Solutions](https://advanced-r-solutions.rbind.io), organized by chapter. It includes excerpts from both books, copied here.

**WARNING, SPOILERS!** If you haven't read Advanced R and intend to complete the quizzes and exercises, don't read this notebook. It contains my (potentially wrong) answers to both.

# 12 Base types

To talk about objects and OOP in R we first need to clear up a fundamental confusion about two uses of the word "object". So far in this book, we've used the word in the general sense captured by John Chambers' pithy quote: "Everything that exists in R is an object". However, while everything _is_ an object, not everything is object-oriented. This confusion arises because the base objects come from S, and were developed before anyone thought that S might need an OOP system. The tools and nomenclature evolved organically over many years without a single guiding principle.

Most of the time, the distinction between objects and object-oriented objects is not important. But here we need to get into the nitty gritty details so we'll use the terms __base objects__ and __OO objects__ to distinguish them.

# 13 S3

S3 is R's first and simplest OO system. S3 is informal and ad hoc, but there is a certain elegance in its minimalism: you can't take away any part of it and still have a useful OO system. For these reasons, you should use it, unless you have a compelling reason to do otherwise. S3 is the only OO system used in the base and stats packages, and it's the most commonly used system in CRAN packages.

S3 is very flexible, which means it allows you to do things that are quite ill-advised. If you're coming from a strict environment like Java this will seem pretty frightening, but it gives R programmers a tremendous amount of freedom.  It may be very difficult to prevent people from doing something you don't want them to do, but your users will never be held back because there is something you haven't implemented yet. Since S3 has few built-in constraints, the key to its successful use is applying the constraints yourself. This chapter will therefore teach you the conventions you should (almost) always follow.

The goal of this chapter is to show you how the S3 system works, not how to use it effectively to create new classes and generics. I'd recommend coupling the theoretical knowledge from this chapter with the practical knowledge encoded in the [vctrs package](https://vctrs.r-lib.org).

## 13.2.1 Exercises

1.  Describe the difference between `t.test()` and `t.data.frame()`.
    When is each function called?

```{r}
ftype(t.test)
ftype(t.data.frame)
s3_dispatch(t.test(formula()))
s3_dispatch(t(data.frame()))
```

Answer: As noted by `sloop::ftype()` and the docs, `t.test()` is a S3 generic, and `t.data.frame()` is an S3 method for `t()` (transpose). `t.data.frame()` is called as a method when calling `t(x)` when `x` is a `data.frame`. `t.test()` calls either `t.test.default` or `t.test.formula`.

AR Solutions: Because of S3's `generic.class()` naming scheme, both functions may initially look similar, while they are in fact unrelated.

- `t.test()` is a *generic* function that performs a t-test. 
- `t.data.frame()` is a *method* that gets called by the generic `t()` to transpose data frame input.

Due to R's S3 dispatch rules, `t.test()` would also get called when `t()` is applied to an object of class `test`

---

2.  Make a list of commonly used base R functions that contain `.` in their
    name but are not S3 methods.

```{r}
ftype(as.character)
ftype(as.data.frame)
ftype(data.frame)
ftype(eval.parent)
ftype(file.path)
ftype(file.copy)
ftype(is.null)
ftype(is.data.frame)
ftype(Sys.localeconv)
ftype(Sys.time)
```

Answer:

- All of the `as.` functions
- `data.frame`
- All of the `file.` functions
- All of the `is.` functions
- All of the `Sys.` functions

AR Solutions: In recent years "snake_case"-style has become increasingly common when naming functions and variables in R. But many functions in base R will continue to be "point.separated", which is why some inconsistency in your R code most likely cannot be avoided. (`install.packages(), read.csv(), list.files(), download.file(), data.frame(), as.character(), Sys.Date(), all.equal(), do.call(), on.exit()`)

---

3.  What does the `as.data.frame.data.frame()` method do? Why is
    it confusing? How could you avoid this confusion in your own
    code?

```{r}
s3_dispatch(as.data.frame(data.frame()))
```

Answer: `as.data.frame.data.frame()` is the method used to coerce a `data.frame` to a `data.frame`. This is confusing because the class contains a period (as does the function call); avoiding periods improves readability: `as_dataframe.dataframe` makes the generic and method clear.

AR Solutions: The function `as.data.frame.data.frame()` implements the `data.frame()` *method* for the `as.data.frame()` *generic*, which coerces objects to data frames.

The name is confusing, because it does not clearly communicate the type of the function, which could be a regular function, a generic or a method. Even if we assume a method, the amount of `.`'s makes it difficult to separate the generic- and the class-part of the name. Is it the `data.frame.data.frame()` method for the `as()` generic? Is it the `frame.data.frame()` method for the `as.data()` generic?

We could avoid this confusion by applying a different naming convention (e.g. "snake_case") for our class and function names.

---

4.  Describe the difference in behaviour in these two calls.

```{r}
set.seed(1014)
some_days <- as.Date("2017-01-31") + sample(10, 5)
mean(some_days)
mean(unclass(some_days))

s3_dispatch(mean(some_days))
s3_dispatch(mean(unclass(some_days)))
```

Answer: the first call calculates the mean using `mean.Date()`, and so returns a date. using `unclass()` changes the date to its underlying value (double) which calculates the mean using `mean.default()`.

AR Solutions: `mean()` is a generic function, which will select the appropriate method based on the class of the input. `some_days` has the class `Date` and `mean.Date(some_days)` will be used to calculate the mean date of `some_days`.

After `unclass()` has removed the class attribute from `some_date`, the default method is chosen. `mean.default(unclass(some_days))` then calculates the mean of the underlying double.

---

5.  What class of object does the following code return? What base type is it 
    built on? What attributes does it use?

```{r}
x <- ecdf(rpois(100, 10))
x

str(x)
```

Answer: the code returns an object of class `ecdf`, which is build on the `stepfun` object, and the `function` base type. It additionally includes the `call` attribute.

AR Solutions: It returns an object of the class `ecdf` (empirical cumulative distribution function) with the superclasses `stepfun` and `function`. The `ecdf` object is built on the base type `closure` (a function). The expression, which was used to create it (`rpois(100, 10)`), is stored in the `call` attribute.

```{r}
typeof(x)
```

---

6.  What class of object does the following code return? What base type is it 
    built on? What attributes does it use?

```{r}
x <- table(rpois(100, 5))
x

str(x)
str(unclass(x))
```

Answer: per the docs, the code returns an object of class `table`, which is built on `array`, which uses the `dimnames` attribute in combination with a vector.

AR Solutions: This code returns a `table` object, which is built upon the `integer` type. The attribute `dimnames` is used to name the elements of the integer vector.

---

## 13.3.4 Exercises

1.  Write a constructor for `data.frame` objects. What base type is a data
    frame built on? What attributes does it use? What are the restrictions
    placed on the individual elements? What about the names?

```{r}
dput(data.frame())
dput(data.frame(a = 1:2, b = 3:4))
unclass(data.frame(a = 1:2, b = 3:4))
```

Answer: code below. The data frame is built on the list base type, and includes the names, row.names, and class attributes. This implementation requires values to be a list, names to be a character, and row.names to be an integer.

```{r}
new_data.frame <- function(values = list(),
                           names = character(length(values)),
                           row.names = integer(length(values))) {
  stopifnot(
    is.list(values),
    is.character(names),
    is.integer(row.names)
  )
  structure(values, names = names, row.names = row.names, class = "data.frame")
}

new_data.frame()
new_data.frame(list(1:3, 4:6, 7:9))
new_data.frame(list(1:3, 4:6, 7:9), names = c("a", "b", "c"), row.names = 1:3)
```

AR Solutions: Data frames are built on named lists of vectors, which all have the same length. Besides the `class` and the column names (`names`), the `row.names` are their only further attribute. This must be a character vector with the same length as the other vectors.

We need to provide the number of rows as an input to make it possible to create data frames with 0 columns but multiple rows.

This leads to the following constructor:

```{r, error = TRUE}
new_data.frame <- function(x, n, row.names = NULL) {
  # Check if the underlying object is a list
  stopifnot(is.list(x))

  # Check all inputs are the same length
  # (This check also allows that x has length 0)
  stopifnot(all(lengths(x) == n))

  if (is.null(row.names)) {
    # Use special row names helper from base R
    row.names <- .set_row_names(n)
  } else {
    # Otherwise check that they're a character vector with the
    # correct length
    stopifnot(is.character(row.names), length(row.names) == n)
  }

  structure(
    x,
    class = "data.frame",
    row.names = row.names
  )
}

# Test
x <- list(a = 1, b = 2)
new_data.frame(x, n = 1)
new_data.frame(x, n = 1, row.names = "l1")

# Create a data frame with 0 columns and 2 rows
new_data.frame(list(), n = 2)
```

Note: AR Solutions approach also validates that all inputs are same length, and supports creation of dataframes with 0 columns but multiple rows.

---

2.  Enhance my `factor()` helper to have better behaviour when one or
    more `values` is not found in `levels`. What does `base::factor()` do
    in this situation?

```{r}
new_factor <- function(x = integer(), levels = character()) {
  stopifnot(
    is.integer(x),
    is.character(levels)
  )

  structure(
    x,
    levels = levels,
    class = "factor"
  )
}
```

Answer: the fix is to update `validate_factor()` to allow `NA` values, since the helper already fills in `NA` when `values` is not found in `levels`. This matches the behavior of `base::factor()`. 

```{r}
validate_factor <- function(x) {
  values <- unclass(x)
  levels <- attr(x, "levels")

  if (!all(is.na(values) | values > 0)) {
    stop(
      "All non-missing `x` values must be greater than zero",
      call. = FALSE
    )
  }

  if (length(levels) < max(values, na.rm = TRUE)) {
    stop(
      "There must be at least as many `levels` as possible values in `x`",
      call. = FALSE
    )
  }

  x
}

my_factor <- function(x = character(), levels = unique(x)) {
  ind <- match(x, levels)
  validate_factor(new_factor(ind, levels))
}

factor(c("a", "a", "b"), levels = "a")
my_factor(c("a", "a", "b"), levels = "a")
```

AR Solutions: `base::factor()` converts these values (silently) into `NA`s.

The `factor()` helper including the constructor (`new_factor()`) and its validator (`validate_factor()`) were given in *Advanced R*. However, as the goal of this question is to throw an early error within the helper, we only repeat the code for the helper:

To improve the `factor()` helper we choose to return an informative error message instead.

```{r, error = TRUE}
factor2 <- function(x, levels = unique(x)) {
  new_levels <- match(x, levels)

  # Error if levels don't include all values
  missing <- unique(setdiff(x, levels))
  if (length(missing) > 0) {
    stop(
      "The following values do not occur in the levels of x: ",
      paste0("'", missing, "'", collapse = ", "), ".",
      call. = FALSE
    )
  }

  validate_factor(new_factor(new_levels, levels))
}

# Test
try(factor2(c("a", "b", "c"), levels = c("a", "b")))
```

---

3.  Carefully read the source code of `factor()`. What does it do that
    my constructor does not?

```{r}
factor(c("a", "a", "b", "c"), labels = c("alpha", "beta", "beta"))
class(factor(1:10, ordered = TRUE))
```

Answer: the base implementation:

- sets the value to `character(0)` if the value is null
- retains value names
- allows an upper bound on the number of levels, `nmax`
- coerces the value to character
- provides a method for excluding values from levels, `exclude`
- provides `labels` for remapping factors
- adds an "ordered" class if the value is ordered (`ordered = TRUE`)

Note: labels are an interesting and unexpected feature of `factor()`

```{r}
factor
```

AR Solutions: The original implementation (`base::factor()`) allows more flexible input for `x`. It coerces `x` to character or replaces it with `character(0)` (in case of `NULL`). It also ensures that the `levels` are unique. This is achieved by setting them via `base::levels<-`, which fails when duplicate values are supplied.

Note: I missed the fact that `base::levels<-` fails when duplicate values are supplied.

---
    
4.  Factors have an optional "contrasts" attribute. Read the help for `C()`, 
    and briefly describe the purpose of the attribute. What type should it 
    have? Rewrite the `new_factor()` constructor to include this attribute.

Answer: per the "contrast {stats}" documentation, contrast matrices are used in fitting analysis of variance and regression models, so the attribute should be a matrix.

```{r}
new_factor <- function(x = integer(), levels = character(), contr = matrix()) {
  stopifnot(is.integer(x), is.character(levels), is.matrix(contr))

  structure(
    x,
    levels = levels,
    class = "factor",
    contrasts = contr
  )
}
```

AR Solutions: When factor variables (representing nominal or ordinal information) are used in statistical models, they are typically encoded as dummy variables and by default each level is compared with the first factor level. However, many different encodings ("contrasts") are possible, see [Contrast](https://en.wikipedia.org/wiki/Contrast_(statistics)).

Within R's formula interface you can wrap a factor in `stats::C()` and specify the contrast of your choice. Alternatively, you can set the `contrasts` attribute of your factor variable, which accepts matrix input. (See `?contr.helmert` or similar for details.)

Our updated `new_factor()` constructor gets a `contrasts` argument, which accepts a numeric matrix or `NULL` (default).

```{r}
# Updated new_factor() constructor
new_factor <- function(x = integer(),
                       levels = character(),
                       contrasts = NULL) {
  stopifnot(is.integer(x))
  stopifnot(is.character(levels))

  if (!is.null(constrasts)) {
    stopifnot(is.matrix(contrasts) && is.numeric(contrasts)) # nolint: conjunct_test_linter.
  }

  structure(
    x,
    levels = levels,
    class = "factor",
    contrasts = contrasts
  )
}
```

---

5.  Read the documentation for `utils::as.roman()`. How would you write a
    constructor for this class? Does it need a validator? What might a helper 
    do?

```{r}
dput(as.roman(3899))

new_roman <- function(x = integer()) {
  stopifnot(is.integer(x))

  structure(
    x,
    class = "roman"
  )
}

new_roman(2022L)
```

Answer: the structure of the class "roman" is simple, an integer with a defined class. A simple constructor would take an integer and return an object of class "roman" as above. A validator is probably not needed, but could validate that the integer is within the supported range of integers (1-3899). A helper might coerce the number to be an integer using `as.integer()`.

AR Solutions: This function transforms numeric input into Roman numbers. It is built on the integer type, which results in the following constructor.

```{r}
new_roman <- function(x = integer()) {
  stopifnot(is.integer(x))
  structure(x, class = "roman")
}
```

The documentation tells us, that only values between 1 and 3899 are uniquely represented, which we then include in our validation function.

```{r}
validate_roman <- function(x) {
  values <- unclass(x)

  if (any(values < 1 | values > 3899)) {
    stop(
      "Roman numbers must fall between 1 and 3899.",
      call. = FALSE
    )
  }

  x
}
```

For convenience, we allow the user to also pass real values to a helper function.

```{r, error = TRUE}
roman <- function(x = integer()) {
  x <- as.integer(x)

  validate_roman(new_roman(x))
}

# Test
roman(c(1, 753, 2019))
try(roman(0))
```

---

## 13.4.4 Exercises

1.  Read the source code for `t()` and `t.test()` and confirm that
    `t.test()` is an S3 generic and not an S3 method. What happens if
    you create an object with class `test` and call `t()` with it? Why?

```{r}
x <- structure(1:10, class = "test")
t(x)
```

Answer: both `t()` and `t.test()` simply call `UseMethod()`, and are S3 generics.

```{r}
t
t.test
```

Creating an object with class `test` and calling `t()` uses the default method since `t.test()` is not a registered method for `t()`, as this code shows:

```{r}
methods("t")
s3_dispatch(t(x))
```

AR Solutions: We can see that `t.test()` is a generic because it calls `UseMethod()`.

```{r}
# or simply call
ftype(t.test)
```

Interestingly, R also provides helpers, which list functions that look like methods, but in fact are not:

```{r}
tools::nonS3methods("stats")
```

When we create an object with class `test`, `t()` dispatches to the `t.default()` method. This happens, because `UseMethod()` simply searches for functions named `paste0("generic", ".", c(class(x), "default"))`.

However, in older versions of R (pre R 4.0.0; when *Advanced R* was written) this behaviour was slightly different. Instead of dispatching to the `t.default()` method, the `t.test()` generic was erroneously treated as a method of `t()` which then dispatched to `t.test.default()` or (when defined) to `t.test.test()`.

---

2.  What generics does the `table` class have methods for?

Answer: `s3_methods_class()` answers this question:

```{r warning = FALSE}
s3_methods_class("table")
```

AR Solutions: This is a simple application of `sloop::s3_methods_class()`.

Interestingly, the `table` class has a number of methods designed to help plotting with base graphics.

```{r}
x <- rpois(100, 5)
plot(table(x))
```

---

3.  What generics does the `ecdf` class have methods for?

Answer:

```{r warning = FALSE}
s3_methods_class("ecdf")
```

AR Solutions: We use the same approach as above.

The methods are primarily designed for display (`plot()`, `print()`, `summary()`), but you can also extract quantiles with `quantile()`.

---

4.  Which base generic has the greatest number of defined methods?

Answer: using code from 6.2.5, identify generics and count defined methods:

```{r, warning = FALSE}
# from 6.2.5 exercises, this code makes a list of all functions in the base package
funs <- Filter(is.function, mget(ls("package:base", all.names = TRUE), inherits = TRUE))

get_method_count <- function(fname) {
  data.frame(
    name = fname,
    # this code should work but doesn't:
    # method_count = nrow(s3_methods_generic(fname))
    method_count = length(methods(fname))
  )
}

map_dfr(names(funs), get_method_count) |>
  arrange(desc(method_count))
```

This brute-force code unsurprisingly identifies `print()` as having the greatest number of methods. A bug in `s3_methods_generic()` prevents its use (error below).

```
Error in gsub(paste0("^", generic_esc, "[.,]"), "", info$method) : 
invalid regular expression '^([.,]', reason 'Missing ')''
```

AR Solutions: A little experimentation (and thinking about the most popular functions) suggests that the `print()` generic has the most defined methods.

Let's verify this programmatically with the tools we have learned in this and the previous chapters.

```{r, message = FALSE}
ls(all.names = TRUE, envir = baseenv()) %>%
  mget(envir = baseenv()) %>%
  keep(is_function) %>%
  names() %>%
  keep(is_s3_generic) %>%
  map(~ set_names(nrow(s3_methods_generic(.x)), .x)) %>%
  flatten_int() %>%
  sort(decreasing = TRUE) %>%
  head()
```

---

5.  Carefully read the documentation for `UseMethod()` and explain why the
    following code returns the results that it does. What two usual rules
    of function evaluation does `UseMethod()` violate?

```{r}
g <- function(x) {
  x <- 10
  y <- 10
  UseMethod("g")
}
g.default <- function(x) c(x = x, y = y)
x <- 1
y <- 1
g(x)
```

Answer: the documentation states in **Technical Details**:

> `UseMethod` creates a new function call with arguments matched as they came in to the generic. Any local variables defined before the call to `UseMethod` are retained (unlike S).

So, `UseMethod()` creates a new function call where `x` matches `x <- 1` and the local variable `y <- 10` is retained. This is passed to `g.default()`. This violates both lazy evaluation and lexical scoping.

AR Solutions: Let's take this step by step. If you call `g.default(x)` directly you get `c(1, 1)` as you might expect. 

The value bound to `x` comes from the argument, the value from `y` comes from the global environment.

But when we call `g(x)` we get `c(1, 10)`.

This is seemingly inconsistent: why does `x` come from the value defined inside of `g()`, and `y` still come from the global environment? It's because `UseMethod()` calls `g.default()` in a special way so that variables defined inside the generic are available to methods. The exception are arguments supplied to the function: they are passed on as is and cannot be affected by code inside the generic.

---

6.  What are the arguments to `[`? Why is this a hard question to answer?

Answer: the possible arguments include the `x`, the object being extracted (or replaced), indices (`i`, `j`, `...`), `name`, `drop`, `exact`, and `value`.

This is a hard question to answer since `[` is an irregular primitive function that can be called in multiple ways, none of which support using all possible arguments.

AR Solutions: The subsetting operator `[` is a primitive and a generic function, which can be confirmed via `ftype()`.

```{r}
ftype(`[`)
```

For primitive functions `formals([)` returns `NULL` so we need to find another way to determine the functions arguments. One possible way to figure out `[`'s arguments would be to inspect the underlying C source code, which can be searched for via `pryr::show_c_source(.Primitive("["))`.
When we inspect the arguments of some of `[`'s methods, we see that the arguments vary with the class of `x`.

```{r}
names(formals(`[.data.frame`))
names(formals(`[.table`))
names(formals(`[.Date`))
names(formals(`[.AsIs`))
```

To finally get a better overview, we have to put in a little more effort and also use `s3_methods_generic()` again.

```{r, message = FALSE, warning = FALSE}
s3_methods_generic("[") %>%
  filter(visible) %>%
  mutate(
    method = paste0("[.", class),
    argnames = purrr::map(method, ~ names(formals(.x))),
    args = purrr::map(method, ~ formals(.x)),
    args = purrr::map2(
      argnames, args,
      ~ paste(.x, .y, sep = " = ")
    ),
    args = purrr::set_names(args, method)
  ) %>%
  pull(args) %>%
  head()
```

---

## 13.5.1 Exercises

1.  Categorise the objects returned by `lm()`, `factor()`, `table()`, 
    `as.Date()`, `as.POSIXct()`, `ecdf()`, `ordered()`, `I()` into the
    styles described above.

Answer:

- `lm()`, `ecdf()`: scalar
- `factor()`, `as.Date()`, `as.POSIXct()`, `ordered`: vector
- `table()`: data frame
- `I()`: same as the source object

AR Solutions: We can categorise the return values into the various object styles by observing how the [number of observations](https://vctrs.r-lib.org/articles/type-size.html#size) is calculated: For vector style classes, `length(x)` represents the number of observations. Record style objects use a list of equal length elements to represent individual components. For data frames and matrices, the observations are represented by the rows. Scalar style objects use a list to represent a single thing.

This leads us to:

- Vector object-style: `factor()`, `table()`, `as.Date()`, `as.POSIXct()`, `ordered()`
- Record object-style: not observed
- Data frame object-style: not observed
- Scalar object-style: `lm()`, `ecdf()`  

The object style of `I()` depends on the input since this function returns a "copy of the object with class `AsIs` prepended to the class(es)".

---

2.  What would a constructor function for `lm` objects, `new_lm()`, look like?
    Use `?lm` and experimentation to figure out the required fields and their
    types.

Answer: the only required field is `formula`, provided that the variables listed in the formula exist (and are both vectors). `formula` must be of type "`formula`".

```{r}
mpg <- pull(mtcars, mpg)
wt <- pull(mtcars, wt)
lm(mpg ~ wt)
```

AR Solutions: The constructor needs to populate the attributes of an `lm` object and check their types for correctness. Let's start by creating a simple `lm` object and explore its underlying base type and attributes:

```{r}
mod <- lm(cyl ~ ., data = mtcars)
typeof(mod)
attributes(mod)
```

As `mod` is built upon a list, we can simply use `map(mod, typeof)` to find out the base types of its elements. (Additionally, we inspect `?lm`, to learn more about the individual attributes.)

```{r}
map_chr(mod, typeof)
```

Now we should have enough information to write a constructor for new `lm` objects.

```{r}
new_lm <- function(coefficients, residuals, effects, rank, fitted.values, assign,
                   qr, df.residual, xlevels, call, terms, model) {
  stopifnot(
    is.double(coefficients), is.double(residuals),
    is.double(effects), is.integer(rank), is.double(fitted.values),
    is.integer(assign), is.list(qr), is.integer(df.residual),
    is.list(xlevels), is.language(call), is.language(terms),
    is.list(model)
  )

  structure(
    list(
      coefficients = coefficients,
      residuals = residuals,
      effects = effects,
      rank = rank,
      fitted.values = fitted.values,
      assign = assign,
      qr = qr,
      df.residual = df.residual,
      xlevels = xlevels,
      call = call,
      terms = terms,
      model = model
    ),
    class = "lm"
  )
}
```

---

## 13.6.3 Exercises

1.  How does `[.Date` support subclasses? How does it fail to support 
    subclasses?

```{r}
`[.Date`
`.Date`
```

Answer: by passing the object's class to `.Date`, `[.Date` avoids the problem of `[.secret` and can return subclasses of `Date`. However, the `.Date` method can return objects that don't contain the class `Date`.

AR Solutions: `[.Date` calls `.Date` with the result of calling `[` on the parent class, along with `oldClass()`:

`.Date` is kind of like a constructor for date classes, although it doesn't check the input is the correct type:

`oldClass()` is basically the same as `class()`, except that it doesn't return implicit classes, i.e. it's basically `attr(x, "class")` (looking at the C code that's exactly what it does, except that it also handles S4 objects).

As `oldClass()` is "basically" `class()`, we can rewrite `[.Date` to make the implementation more clear:

```{r}                                                            
`[.Date` <- function(x, ..., drop = TRUE) {
  out <- NextMethod("[")
  class(out) <- class(x)
  out
}
```
                                                        
So, `[.Date` ensures that the output has the same class as in the input. But what about other attributes that a subclass might possess? They get lost:

```{r}
x <- structure(1:4, test = "test", class = c("myDate", "Date"))
attributes(x[1])
```

Note: this is a better explanation of how it fails to support subclasses; it drops non-class attributes.

---

2.  R has two classes for representing date time data, `POSIXct` and
    `POSIXlt`, which both inherit from `POSIXt`. Which generics have
    different behaviours for the two classes? Which generics share the same
    behaviour?

Answer: if a generic is defined for either class, it must have different behaviors. Generics that aren't defined for either class must have the same behavior. We can get the list of generics defined for either class using `sloop::s3_methods_class()`. These generics have different behaviors for the two classes:

```{r, warning = FALSE}
ct_generics <- s3_methods_class("POSIXct")$generic
lt_generics <- s3_methods_class("POSIXlt")$generic
defined_generics <- union(ct_generics, lt_generics)
defined_generics
```

We adapt work from previous exercises to get the list of all generics, and identify S3 generics that aren't defined for either class, which must have the same behavior:

```{r}
s3_generic_funs <- Filter(is_s3_generic, names(funs))
union(setdiff(defined_generics, s3_generic_funs), setdiff(s3_generic_funs, defined_generics))
```

Note: this code doesn't quite work, since `Summary` is a generic according to `isGeneric` and is not according to `is_s3_generic`, and `weighted.mean` is a non-base S3 generic.

AR Solutions: To answer this question, we have to get the respective generics

```{r, warning = FALSE}
generics_t <- s3_methods_class("POSIXt")$generic
generics_ct <- s3_methods_class("POSIXct")$generic
generics_lt <- s3_methods_class("POSIXlt")$generic
```

The generics in `generics_t` with a method for the superclass `POSIXt` potentially share the same behaviour for both subclasses. However, if a generic has a specific method for one of the subclasses, it has to be subtracted:

```{r}
# These generics provide subclass-specific methods
union(generics_ct, generics_lt)

# These generics share (inherited) methods for both subclasses
setdiff(generics_t, union(generics_ct, generics_lt))
```

Note: again, advanced R gives a better answer to this question, by comparing to the generics supported by the parent class.

---

3.  What do you expect this code to return? What does it actually return?
    Why?

```{r}
generic2 <- function(x) UseMethod("generic2")
generic2.a1 <- function(x) "a1"
generic2.a2 <- function(x) "a2"
generic2.b <- function(x) {
  class(x) <- "a1"
  NextMethod()
}
generic2(structure(list(), class = c("b", "a2")))
```

Answer: I'd expect this code to return `"a2"` with a class of `c("a1", "a2")`. It actually returns "a2" with a class of `character`. This is because when `generic2.b` calls `NextMethod()`, `generic2.a2` is called which returns a bare `"a2"`, dropping the class assigned by `generic2.b`.

```{r}
class(generic2(structure(list(), class = c("b", "a2"))))
```


AR Solutions: When we execute the code above, this is what is happening:

- we pass an object of classes `b` and `a2` to `generic2()`, which prompts R to look for a method`generic2.b()`
- the method `generic2.b()` then changes the class to `a1` and calls `NextMethod()`
- One would think that this will lead R to call `generic2.a1()`, but in fact, as mentioned in *Advanced R*, `NextMethod()` 
    > doesn’t actually work with the class attribute of the object, but instead uses a special global variable (.Class) to keep track of which method to call next.
    
   This is why `generic2.a2()` is called instead.

```{r}
generic2(structure(list(), class = c("b", "a2")))
```

Let's just double check the statement above and evaluate `.Class` explicitly within the `generic2.b()` method.

```{r}
generic2.b <- function(x) {
  class(x) <- "a1"
  print(.Class)
  NextMethod()
}

generic2(structure(list(), class = c("b", "a2")))
```

---

## 13.7.5 Exercises

1.  Explain the differences in dispatch below:

```{r}
length.integer <- function(x) 10

x1 <- 1:5
class(x1)
s3_dispatch(length(x1))

x2 <- structure(x1, class = "integer")
class(x2)
s3_dispatch(length(x2))
```

Answer: `x1` dispatches using implicit class, which is `c("integer", "numeric")`. `x2` has an explicit class, so it is dispatched to `length.integer`.

```{r}
s3_class(x1)
s3_class(x2)
```

AR Solutions: `class()` returns `integer` in both cases. However, while the class of `x1` is created implicitly and inherits from the `numeric` class, the class of `x2` is set explicitly. This is important because `length()` is an internal generic and internal generics only dispatch to methods when the class attribute has been set, i.e. internal generics do not use implicit classes.

An object has no explicit class if `attr(x, "class")` returns `NULL`:

```{r}
attr(x1, "class")
attr(x2, "class")
```

To see the relevant classes for the S3 dispatch, one can use `sloop::s3_class()`:

```{r}
s3_class(x1) # implicit

s3_class(x2) # explicit
```

For a better understanding of `s3_dipatch()`'s output we quote from `?s3_dispatch`:
- => method exists and is found by `UseMethod()`.
- -> method exists and is used by `NextMethod()`.
- \* method exists but is not used.
- Nothing (and greyed out in console): method does not exist.

Note: AR Solutions includes an additional important detail, that "internal generics only dispatch to methods when the class attribute has been set", which explains why the internal generic is used for `x1`.

---

2.  What classes have a method for the `Math` group generic in base R? Read
    the source code. How do the methods work?

Answer: four base classes have a method for `Math`: `data.frame`, `Date`, `difftime`, `factor`, and `POSIXt`.

```{r}
s3_methods_generic("Math") |>
  filter(source == "base")
```

```{r}
Math.data.frame
Math.Date
Math.difftime
Math.factor
Math.POSIXt
```

- `Math.data.frame` performs math if all columns (variables) are either numeric, logical or complex,
  and errors if any are not
- `Math.difftime` performs only certain math functions: `abs`, `sign`, `floor`, `ceiling`, `trunc`,
  `round`, and `signif`, and errors for other functions
- `Math.Date`, `Math.factor`, and `Math.POSIXt` always error

AR Solutions: : The following functions belong to this group (see ?`Math`):

- `abs`, `sign`, `sqrt`, `floor`, `ceiling`, `trunc`, `round`, `signif`
- `exp`, `log`, `expm1`, `log1p`, `cos`, `sin`, `tan`, `cospi`, `sinpi`, `tanpi`, `acos`, `asin`,
  `atan`, `cosh`, `sinh`, `tanh`, `acosh`, `asinh`, `atanh`
- `lgamma`, `gamma`, `digamma`, `trigamma`
- `cumsum`, `cumprod`, `cummax`, `cummin`

The following classes have a method for this group generic:

```{r}
s3_methods_generic("Math")
```

To explain the basic idea, we just overwrite the data frame method:

```{r}
Math.data.frame <- function(x) "hello"
```

Now all functions from the math generic group, will return `"hello"`

```{r}
abs(mtcars)
exp(mtcars)
lgamma(mtcars)
```

Of course, different functions should perform different calculations. Here `.Generic` comes into play, which provides us with the calling generic as a string

```{r}
Math.data.frame <- function(x, ...) {
  .Generic
}
abs(mtcars)
exp(mtcars)
lgamma(mtcars)
rm(Math.data.frame)
```

The original source code of `Math.data.frame()` is a good example on how to invoke the string returned by `.Generic` into a specific method. `Math.factor()` is a good example of a method, which is simply defined for better error messages.

Note: I prefer my answer to the AR Solutions answer.

---

3.  `Math.difftime()` is more complicated than I described. Why?

Answer: the explanation in the text omits the fact that `Math.difftime()` first checks that the math function being called.

AR Solutions: `Math.difftime()` also excludes cases apart from `abs`, `sign`, `floor`, `ceiling`, `trunc`, `round` and `signif` and needs to return a fitting error message.

---

# 14 R6

```{r}
rm(list = ls())
```

This chapter describes the R6 OOP system. R6 has two special properties:

* It uses the encapsulated OOP paradigm, which means that methods belong to 
  objects, not generics, and you call them like `object$method()`. 
  
* R6 objects are __mutable__, which means that they are modified in place, and 
  hence have reference semantics.

If you've learned OOP in another programming language, it's likely that R6 will feel very natural, and you'll be inclined to prefer it over S3. Resist the temptation to follow the path of least resistance: in most cases R6 will lead you to non-idiomatic R code. We'll come back to this theme in Section 16.3.

R6 is very similar to a base OOP system called __reference classes__, or RC for short. I describe why I teach R6 and not RC in Section 14.5.

## 14.2.6 Exercises

1.  Create a bank account R6 class that stores a balance and allows you to 
    deposit and withdraw money. Create a subclass that throws an error 
    if you attempt to go into overdraft. Create another subclass that allows
    you to go into overdraft, but charges you a fee.

Answer: `Account` stores a balance as an integer value of cents, and prints in dollars (to avoid problems with floating point math).

```{r}
Account <- R6Class("Account", list(
  balance = 0L,
  initialize = function(balance = 0L) {
    stopifnot(is.integer(balance), length(balance) == 1)
    self$balance <- balance
  },
  print = function(...) {
    cat("Account: \n")
    cat("  Balance: $", format(self$balance / 100, nsmall = 2), "\n", sep = "")
    invisible(self)
  },
  deposit = function(value) {
    stopifnot(is.integer(value), length(value) == 1)
    self$balance <- self$balance + value
    invisible(self)
  },
  withdraw = function(value) {
    stopifnot(is.integer(value), length(value) == 1)
    self$balance <- self$balance - value
    invisible(self)
  }
))

AccountOverdraft <- R6Class("AccountOverdraft",
  inherit = Account,
  public = list(
    withdraw = function(value) {
      stopifnot(value <= self$balance)
      super$withdraw(value)
    }
  )
)

AccountOverdraftFee <- R6Class("AccountOverdraftFee",
  inherit = Account,
  public = list(
    withdraw = function(value) {
      if (value > self$balance) value <- value + 500L
      super$withdraw(value)
    }
  )
)

try(Account$new(1.5))

a <- Account$new(100L)
a$deposit(100L)
a$withdraw(250L)
a

a2 <- AccountOverdraft$new(200L)
try(a2$withdraw(250L))
a2$withdraw(150L)
a2

a3 <- AccountOverdraftFee$new(200L)
a3$withdraw(250L)
a3
```

AR Solutions: `BankAccount` is a similar, simpler approach:

```{r}
BankAccount <- R6Class(
  classname = "BankAccount",
  public = list(
    balance = 0,
    deposit = function(dep = 0) {
      self$balance <- self$balance + dep
      invisible(self)
    },
    withdraw = function(draw) {
      self$balance <- self$balance - draw
      invisible(self)
    }
  )
)

BankAccountStrict <- R6Class(
  classname = "BankAccountStrict",
  inherit = BankAccount,
  public = list(
    withdraw = function(draw = 0) {
      if (self$balance - draw < 0) {
        stop("Your `withdraw` must be smaller ",
          "than your `balance`.",
          call. = FALSE
        )
      }
      super$withdraw(draw = draw)
    }
  )
)

BankAccountCharging <- R6Class(
  classname = "BankAccountCharging",
  inherit = BankAccount,
  public = list(
    withdraw = function(draw = 0) {
      if (self$balance - draw < 0) {
        draw <- draw + 1
      }
      super$withdraw(draw = draw)
    }
  )
)
```

Note: my object name, `AccountOverdraftFee` is misleading, since it is not a subclass of `AccountOverdraft` - the R Solutions naming convention is clearer.

---

2.  Create an R6 class that represents a shuffled deck of cards. You should be
    able to draw cards from the deck with `$draw(n)`, and return all cards to 
    the deck and reshuffle with `$reshuffle()`. Use the following code to make 
    a vector of cards.

```{r}
suit <- c("♠", "♥", "♦", "♣")
value <- c("A", 2:10, "J", "Q", "K")
cards <- paste0(rep(value, 4), suit)
```

Answer: `Cards` shuffles the deck using `sample()`. The `draw()` function prints the cards drawn and moves them into play.

```{r}
Cards <- R6Class("Cards", list(
  deck = sample(cards, 52),
  play = NULL,
  print = function(...) {
    cat("Cards: \n")
    cat("  Deck:", self$deck, "\n", sep = " ")
    cat("  Play:", self$play, "\n", sep = " ")
    invisible(self)
  },
  draw = function(n = 1L) {
    stopifnot(is.integer(n), length(n) == 1, n > 0, n <= length(self$deck))
    d <- head(self$deck, n)
    self$play <- c(self$play, d)
    self$deck <- tail(self$deck, -n)
    invisible(self)
  },
  reshuffle = function() {
    self$deck <- sample(cards, 52)
    self$play <- NULL
    invisible(self)
  }
))

d <- Cards$new()
d
try(d$draw(1.5))
try(d$draw(0L))
try(d$draw(53L))
d$draw(13L)
d
d$draw(13L)
d
d$reshuffle()
d
```

AR Solutions: Our new `ShuffledDeck` class will use `sample()` and positive integer subsetting to implement the reshuffling and drawing functionality. We also add a check, so you cannot draw more cards than there are left in the deck.

```{r}
ShuffledDeck <- R6Class(
  classname = "ShuffledDeck",
  public = list(
    deck = NULL,
    initialize = function(deck = cards) {
      self$deck <- sample(deck)
    },
    reshuffle = function() {
      self$deck <- sample(cards)
      invisible(self)
    },
    n = function() {
      length(self$deck)
    },
    draw = function(n = 1) {
      if (n > self$n()) {
        stop("Only ", self$n(), " cards remaining.", call. = FALSE)
      }

      output <- self$deck[seq_len(n)]
      self$deck <- self$deck[-seq_len(n)]
      output
    }
  )
)
```

Notes: it seems simpler to set `deck` directly instead of using an `initialize` function, although `sample(cards)` is cleaner. `self$n()` is a useful function, and the choice of removing and returning cards with `draw()` is probably a better design. I prefer my approach using `head()` and `tail()` over R Solutions use of `seq_len()`.

Bug: per 14.4.3 setting fields directly fixes the order when the class is *defined*. This results in a bug: `Cards` has the same starting shuffle for each new object, but `ShuffledDeck` does not:

```{r}
Cards$new()
Cards$new()
ShuffledDeck$new()$draw(52)
ShuffledDeck$new()$draw(52)
```

---

3.  Why can't you model a bank account or a deck of cards with an S3 class?

Answer: because S3 classes are copied-on-modify, they lack state. R6 classes are stateful since they are modified in-place. Both bank accounts and a deck of cards are stateful objects.

AR Solutions: Because S3 classes obey R's usual semantics of copy-on-modify: every time you deposit money into your bank account or draw a card from the deck, you'd get a new copy of the object.

It is possible to combine S3 classes with an environment (which is how R6 works), but it is ill-advised to create an object that looks like a regular R object but has reference semantics.

---

4.  Create an R6 class that allows you to get and set the current time zone. 
    You can access the current time zone with `Sys.timezone()` and set it 
    with `Sys.setenv(TZ = "newtimezone")`. When setting the time zone, make
    sure the new time zone is in the list provided by `OlsonNames()`.

Answer: `TimeZone` is initialized with the current time zone, stored as `tz` in the object.

```{r}
orig_tz <- Sys.timezone()

TimeZone <- R6Class("TimeZone", list(
  tz = Sys.timezone(),
  get = function() {
    cat("Current time zone: ", self$tz, "\n", sep = "")
    invisible(self)
  },
  set = function(tz) {
    stopifnot(tz %in% OlsonNames())
    Sys.setenv(TZ = tz)
    self$tz <- tz
    cat("Set time zone: ", self$tz, "\n", sep = "")
    invisible(self)
  }
))

tz <- TimeZone$new()
tz$get()
date()
tz$set("America/New_York")
date()

Sys.setenv(TZ = orig_tz)
```

Bug: like `Cards`, `TimeZone` should set `tz` in `initialize()`.

AR Solutions: To create an R6 class that allows us to get and set the time zone, we provide the respective functions as public methods to the R6 class.

```{r}
Timezone <- R6Class(
  classname = "Timezone",
  public = list(
    get = function() {
      Sys.timezone()
    },
    set = function(value) {
      stopifnot(value %in% OlsonNames())
      old <- self$get()
      Sys.setenv(TZ = value)
      invisible(old)
    }
  )
)
```

(When setting, we return the old value invisibly because this makes it easy to restore the previous value.)

---

5.  Create an R6 class that manages the current working directory.
    It should have `$get()` and `$set()` methods.

Answer: similar to `TimeZone`, `WorkingDirectory` initializes `wd` to `getwd()`.

```{r warning = FALSE}
orig_wd <- getwd()

WorkingDirectory <- R6Class("WorkingDirectory", list(
  wd = getwd(),
  get = function() {
    cat("Current directory: ", self$wd, "\n", sep = "")
    invisible(self)
  },
  set = function(dir) {
    self$wd <- setwd(dir)
    cat("Set directory: ", self$wd, "\n", sep = "")
    invisible(self)
  }
))

wd <- WorkingDirectory$new()
wd$get()
wd$set("assets")
list.files()

setwd(orig_wd)
```

AR Solutions: Take a look at the following implementation, which is quite minimalistic:

```{r}
WorkingDirectory <- R6Class(
  classname = "WorkingDirectory",
  public = list(
    get = function() {
      getwd()
    },
    set = function(value) {
      setwd(value)
    }
  )
)
```

Bug: like `Cards`, `WorkingDirectory` should set `wd` in `initialize()`. Also, as AR Solutions notes in a comment:

> You should never do `get = getwd()` etc because in packages, that inlines the function definition at package build time, creating a subtle dependency that will cause bugs that are extremely difficult to track down

---

6.  Why can't you model the time zone or current working directory with an S3
    class?

Answer: per exercise 3, time zone and current working directory are both stateful.

AR Solutions: Because S3 classes are not suitable for modelling a state that changes over time. S3 methods should (almost) always return the same result when called with the same inputs.

---

7.  What base type are R6 objects built on top of? What attributes do they
    have?

Answer: R6n objects are built on environments, and have a class attribute with an S3 hierarchy including the base R6 class, per 14.2.5.

```{r}
typeof(tz)
attributes(tz)
```

AR Solutions: R6 objects are built on top of environments. They have a `class` attribute, which is a character vector containing the class name, the name of any super classes (if existent) and the string `"R6"` as the last element.

---

## 14.3.3 Exercises

1.  Create a bank account class that prevents you from directly setting the 
    account balance, but you can still withdraw from and deposit to. Throw
    an error if you attempt to go into overdraft.

Answer: `SecureAccount` uses a private internal balance exposed through an active function.

```{r}
SecureAccount <- R6Class("SecureAccount",
  private = list(
    .balance = 0L
  ),
  public = list(
    deposit = function(value) {
      stopifnot(is.integer(value), length(value) == 1)
      private$.balance <- private$.balance + value
      invisible(self)
    },
    withdraw = function(value) {
      stopifnot(is.integer(value), length(value) == 1, value <= private$.balance)
      private$.balance <- private$.balance - value
      invisible(self)
    }
  ),
  active = list(
    balance = function(value) {
      if (missing(value)) {
        private$.balance
      } else {
        stop("`$balance` is read only", call. = FALSE)
      }
    }
  )
)

sa <- SecureAccount$new()
sa$deposit(10L)
sa$withdraw(5L)
try(sa$withdraw(20L))
try(sa$balance <- 100L)
sa$balance
```

AR Solutions: To fulfill this requirement, we make balance a private field. The user has to use the `$deposit()` and `$withdraw()` methods which have access to the balance field.

```{r, error = TRUE}
BankAccountStrict2 <- R6Class(
  classname = "BankAccountStrict2",
  public = list(
    deposit = function(dep = 0) {
      private$balance <- private$balance + dep
      invisible(self)
    },
    withdraw = function(draw = 0) {
      if (private$balance - draw < 0) {
        stop(
          "Your `withdraw` must be smaller ",
          "than your `balance`.",
          call. = FALSE
        )
      }
      private$balance <- private$balance - draw
      invisible(self)
    }
  ),
  private = list(
    balance = 0
  )
)
```

---

2.  Create a class with a write-only `$password` field. It should have 
    `$check_password(password)` method that returns `TRUE` or `FALSE`, but 
    there should be no way to view the complete password.

Answer: `Login` implements a write-only `$password` field using an active function. `check_password()` will error if a password hasn't been set. In real life, this should be implemented with a secure password hashing function, like `bcrypt` or `scrypt`.

```{r}
Login <- R6Class("Login",
  private = list(
    .password = NULL
  ),
  public = list(
    check_password = function(password) {
      stopifnot(is.character(password), length(password) == 1)
      if (is.null(private$.password)) stop("Error: password not set")
      password == private$.password
    }
  ),
  active = list(
    password = function(value) {
      if (missing(value)) {
        stop("`$password` is write only", call. = FALSE)
      } else {
        stopifnot(is.character(value), length(value) == 1)
        private$.password <- value
        invisible(self)
      }
    }
  )
)

l <- Login$new()
try(l$check_password("Password1"))
l$password <- "Password1"
l$check_password("Password1")
l$check_password("Password2")
try(l$password)
```

AR Solutions: To protect the password from changes and direct access, the password will be a private field. Further, our `Password` will get its own print method which hides the password.

```{r}
Password <- R6Class(
  classname = "Password",
  public = list(
    print = function(...) {
      cat("<Password>: ********\n")
      invisible(self)
    },
    set = function(value) {
      private$password <- value
    },
    check = function(password) {
      identical(password, private$password)
    }
  ),
  private = list(
    password = NULL
  )
)
```

Let's create one instance of our new class and confirm that the password is neither accessible nor visible, but still check-able.

```{r, error = TRUE}
my_pw <- Password$new()
my_pw$set("snuffles")
my_pw$password
my_pw
my_pw$check("snuggles")
my_pw$check("snuffles")
```

Note: the AR Solutions class properly masks the password when calling `print()`, but mine does not:

```{r}
l
```

---

3.  Extend the `Rando` class with another active binding that allows you to 
    access the previous random value. Ensure that active binding is the only
    way to access the value.

Answer: the updated `Rando` class sets a private value when `random` is run that is accessible only through `previous`, initially set to `NA`.

```{r}
Rando <- R6::R6Class("Rando",
  private = list(
    .previous = NA
  ),
  active = list(
    random = function(value) {
      if (missing(value)) {
        private$.previous <- runif(1)
        private$.previous
      } else {
        stop("Can't set `$random`", call. = FALSE)
      }
    },
    previous = function(value) {
      if (missing(value)) {
        private$.previous
      } else {
        stop("Can't set `$previous`", call. = FALSE)
      }
    }
  )
)
x <- Rando$new()

x$previous
x$random
x$previous
x$random
x$previous
```

AR Solutions: To access the previous random value from an instance, we add a private `$last_random` field to our class, and we modify `$random()` to write to this field, whenever it is called. To access the `$last_random` field we provide `$previous()`.

```{r}
Rando <- R6::R6Class(
  classname = "Rando",
  private = list(
    last_random = NULL
  ),
  active = list(
    random = function(value) {
      if (missing(value)) {
        private$last_random <- runif(1)
        private$last_random
      } else {
        stop("Can't set `$random`.", call. = FALSE)
      }
    },
    previous = function(value) {
      if (missing(value)) {
        private$last_random
      }
    }
  )
)
```

---

4.  Can subclasses access private fields/methods from their parent? Perform
    an experiment to find out.

Answer: `LoginEvil` shows that subclasses *can* access private fields from their parent.

```{r}
LoginEvil <- R6Class("LoginEvil",
  inherit = Login,
  public = list(
    steal_password = function() {
      cat("The password is: ", private$.password, "\n", sep = "")
    }
  )
)

evil <- LoginEvil$new()
evil$password <- "SuperSecretPassword"
evil$steal_password()
```

AR Solutions: To find out if private fields/methods can be accessed from subclasses, we first create a class `A` with a private field `foo` and a private method `bar()`. Afterwards, an instance of a subclass `B`  is created and calls the `foobar()` methods, which tries to access the `foo` field and the `bar()` method from its superclass `A`.

```{r}
A <- R6Class(
  classname = "A",
  private = list(
    field = "foo",
    method = function() {
      "bar"
    }
  )
)
B <- R6Class(
  classname = "B",
  inherit = A,
  public = list(
    test = function() {
      cat("Field:  ", super$field, "\n", sep = "")
      cat("Method: ", super$method(), "\n", sep = "")
    }
  )
)
B$new()$test()
```

We conclude that subclasses can access private methods from their superclasses, but not private fields.

Note: as I discovered, private fields are not accessible through `super$`, but *are* accessible through `private$`.

---

## 14.4.4 Exercises

1.  Create a class that allows you to write a line to a specified file. 
    You should open a connection to the file in `$initialize()`, append a
    line using `cat()` in `$append_line()`, and close the connection in
    `$finalize()`.

Answer: `WriteFile` implemented below.

```{r}
WriteFile <- R6Class("WriteFile", list(
  wfile = NULL,
  initialize = function(filename) {
    self$wfile <- file(filename, "a")
  },
  append_line = function(line) {
    cat(line, file = self$wfile)
  },
  finalize = function() {
    close(self$wfile)
  }
))

tmp <- tempfile()
wf <- WriteFile$new(tmp)
wf$append_line("The quick brown fox jumps over the lazy dog.\n")
readLines(tmp)
```

AR Solutions: Our `FileWriter` class will create a connection to a file at initialization. Therefore, we open a connection to a user specified file during the initialisation. Note that we need to set `open = "a"` in `file()` to open connection for appending text. Otherwise, `cat()` would only work when applied to files, but not with connections as explicitly asked for in the exercise. Further, we add the `append_line()` method and a `close()` statement as finalizer.

```{r, eval = TRUE, error = TRUE}
FileWriter <- R6::R6Class(
  classname = "FileWriter",
  public = list(
    con = NULL,
    initialize = function(filename) {
      self$con <- file(filename, open = "a")
    },
    finalize = function() {
      close(self$con)
    },
    append_line = function(x) {
      cat(x, "\n", sep = "", file = self$con)
    }
  )
)
```

---

# 15 S4

```{r}
rm(list = ls())
```

S4 provides a formal approach to functional OOP. The underlying ideas are similar to S3 (the topic of Chapter 13), but implementation is much stricter and makes use of specialised functions for creating classes (`setClass()`), generics (`setGeneric()`), and methods (`setMethod()`). Additionally, S4 provides both multiple inheritance (i.e. a class can have multiple parents) and multiple dispatch (i.e. method dispatch can use the class of multiple arguments).

An important new component of S4 is the __slot__, a named component of the object that is accessed using the specialised subsetting operator `@` (pronounced at). The set of slots, and their classes, forms an important part of the definition of an S4 class.

## 15.2.1 Exercises

1.  `lubridate::period()` returns an S4 class. What slots does it have?
    What class is each slot? What accessors does it provide?

Answer: the S4 class, `Period`, has 6 slots: .Data, year, month, day, hour, and minute, all numeric. I couldn't figure out how to get the accessors.

```{r}
p <- lubridate::period()
is(p)
vapply(slotNames(p), function(x) class(slot(p, x)), character(1))
```

AR Solutions: Objects of the S4 `Period` class have six slots named `year`, `month`, `day`, `hour`, `minute`, and `.Data` (which contains the number of seconds). All slots are of type double. Most fields can be retrieved by an identically named accessor (e.g. `lubridate::year()` will return the field), use `second()` to get the `.Data` slot.

As a short example, we create a period of 1 second, 2 minutes, 3 hours, 4 days and 5 weeks.

```{r}
example_12345 <- lubridate::period(
  c(1, 2, 3, 4, 5),
  c("second", "minute", "hour", "day", "week")
)
```

This should add up to a period of 39 days, 3 hours, 2 minutes and 1 second.

```{r}
example_12345
```

When we inspect `example_12345`, we see the fields and infer that the seconds are stored in the `.Data` field.

```{r}
str(example_12345)
```

Notes: it is easier to just use `getClass` to enumerate slots and slot classes:

```{r}
getClass(is(lubridate::period()))
```

Methods defined for a particular class can be listed using `showMethods()`, and S4 methods can be listed using `.S4methods()`:

```{r}
.S4methods("Period")
```

`Period` has no S4 methods, only regular functions.

---

2.  What other ways can you find help for a method? Read `?"?"` and
    summarise the details.

Answer: there are three ways to get help for a method:

1. `?function` provides the function documentation
1. `methods?function` provides the overall documentation methods for the function
1. `methods?function(signature)` provides the documentation for a specific method

AR Solutions: Besides adding `?` in front of a function call (i.e. `?method()`), we may find:

- general documentation for a generic via `?genericName`
- general documentation for the methods of a generic via `methods?genericName`
- documentation for a specific method via `ClassName?methodName`.

---

## 15.3.6 Exercises

1.  Extend the Person class with fields to match `utils::person()`.
    Think about what slots you will need, what class each slot should have,
    and what you'll need to check in your validity method.

Answer: the `Person` class uses the fields from `utils::person()`, excluding deprecated fields. Per the docs, `family` must be a string (length 1), others can be vectors (but this is not enforced).

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Person",
  slots = c(
    given = "character",
    family = "character",
    email = "character",
    role = "character",
    comment = "character"
  ),
  prototype = list(
    given = NA_character_,
    family = NA_character_,
    email = NA_character_,
    role = NA_character_,
    comment = NA_character_
  )
)

setValidity("Person", function(object) {
  # regex from https://www.nicebread.de/validating-email-adresses-in-r/
  regex <- "\\<[A-Z0-9._%+-]+@[A-Z0-9.-]+\\.[A-Z]{2,}\\>"

  if (length(object@family) > 1) {
    "@family must be a string (length 1)"
  } else if (!is.na(object@email) & !grepl(regex, object@email, ignore.case = TRUE)) {
    "invalid @email"
  } else {
    TRUE
  }
}) |>
  invisible()

Person <- function(given = NA_character_, family = NA_character_,
                   email = NA_character_, role = NA_character_, comment = NA_character_) {
  given <- as.character(given)
  family <- as.character(family)
  email <- as.character(email)
  role <- as.character(role)
  comment <- as.character(comment)

  new("Person", given = given, family = family, email = email, role = role, comment = comment)
}

Person("John", "Benninghoff", email = "john@example.com", comment = 51)
try(Person("John", c("Three", "Last", "Names")))
try(Person("John", email = "invalid@local"))
```

AR Solutions: The Person class from *Advanced R* contains the slots `name` and `age`. The person class from the `{utils}` package contains the slots `given` (vector of given names), `family`, `role`, `email` and `comment` (see `?utils::person`). 

All slots from `utils::person()` besides `role` must be of type character and length 1. The entries in the `role` slot must match one of the following abbreviations "aut", "com", "cph", "cre", "ctb", "ctr", "dtc", "fnd", "rev", "ths", "trl". Therefore, `role` might be of different length than the other slots and we'll add a corresponding constraint within the validator.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
# Definition of the Person class
setClass("Person",
  slots = c(
    age = "numeric",
    given = "character",
    family = "character",
    role = "character",
    email = "character",
    comment = "character"
  ),
  prototype = list(
    age = NA_real_,
    given = NA_character_,
    family = NA_character_,
    role = NA_character_,
    email = NA_character_,
    comment = NA_character_
  )
)
# Helper to create instances of the Person class
Person <- function(given, family,
                   age = NA_real_,
                   role = NA_character_,
                   email = NA_character_,
                   comment = NA_character_) {
  age <- as.double(age)
  new("Person",
    age = age,
    given = given,
    family = family,
    role = role,
    email = email,
    comment = comment
  )
}
# Validator to ensure that each slot is of length one
setValidity("Person", function(object) { # nolint: cyclocomp_linter.
  invalids <- character(0)
  if (length(object@age) != 1 ||
    length(object@given) != 1 ||
    length(object@family) != 1 ||
    length(object@email) != 1 ||
    length(object@comment) != 1) {
    invalids <- paste0(
      "@name, @age, @given, @family, @email, ",
      "@comment must be of length 1"
    )
  }

  known_roles <- c(
    NA_character_, "aut", "com", "cph", "cre", "ctb",
    "ctr", "dtc", "fnd", "rev", "ths", "trl"
  )

  if (!all(object@role %in% known_roles)) {
    paste(
      "@role(s) must be one of",
      paste(known_roles, collapse = ", ") # nolint: paste_linter.
    )
  }

  if (length(invalids)) {
    return(invalids)
  }
  TRUE
}) |>
  invisible()
```

Notes: while I missed including `age`, our implementations are similar except for validations.

---

2.  What happens if you define a new S4 class that doesn't have any slots?
    (Hint: read about virtual classes in `?setClass`.)

Answer: defining a new class with no slots makes it a virtual class, which can't be used to create new objects, but can be used to define a class union.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Virtual")
try(new("Virtual"))
setClassUnion("Virtual", "Person")
getClass("Virtual")
```

AR Solutions: It depends on the other arguments. If we inherit from another class, we get the same slots. But something interesting happens if we don't inherit from an existing class. We get a virtual class. A virtual class can't be instantiated:

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Human")
try(new("Human"))
```

But can be inherited from:

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Programmer", contains = "Human")
```

---

3.  Imagine you were going to reimplement factors, dates, and data frames in 
    S4. Sketch out the `setClass()` calls that you would use to define the 
    classes. Think about appropriate `slots` and `prototype`.

Answer: sketch below. Default values would be set in `initialize()`.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Factor",
  slots = list(
    x = "character",
    levels = "character",
    labels = "character",
    exclude = "character",
    ordered = "logical",
    nmax = "integer"
  ),
  prototype = list(
    x = character(0),
    levels = character(0),
    labels = character(0),
    exclude = NA_character_,
    ordered = FALSE,
    nmax = NA_integer_
  )
)

setClass("S4Date",
  slots = list(
    days = "integer"
  ),
  prototype = list(
    days = 0L
  )
)

setClass("DataFrame",
  slots = list(
    data = "matrix",
    col_names = "character",
    row_names = "character"
  ),
  prototype = list(
    data = matrix(nrow = 0, ncol = 0),
    col_names = NULL,
    row_names = NULL
  )
)
```

AR Solutions: For all these classes we need one slot for the data and one slot per attribute. Keep in mind, that inheritance matters for ordered factors and dates. For data frames, special checks like equal lengths of the underlying list's elements should be done within a validator.

For simplicity we don't introduce an explicit subclass for ordered factors. Instead, we introduce `ordered` as a slot.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Factor",
  slots = c(
    data = "integer",
    levels = "character",
    ordered = "logical"
  ),
  prototype = list(
    data = integer(),
    levels = character(),
    ordered = FALSE
  )
)
new("Factor", data = c(1L, 2L), levels = letters[1:3])
```

The `Date2` class stores its dates as integers, similarly to base R which uses doubles. Dates don't have any other attributes.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Date2",
  slots = list(
    data = "integer"
  ),
  prototype = list(
    data = integer()
  )
)

new("Date2", data = 1L)
```

Our `DataFrame` class consists of a list and a slot for `row.names`. Most of the logic (e.g. checking that all elements of the list are a vector, and that they all have the same length) would need to be part of a validator.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("DataFrame",
  slots = c(
    data = "list",
    row.names = "character"
  ),
  prototype = list(
    data = list(),
    row.names = character(0)
  )
)

new("DataFrame", data = list(a = 1, b = 2))
```

Notes:

- A key insight from AR Solutions is that the objects need one slot for data and one for each
  attribute
- My implementation of `Factor` is wrong, where the AR Solutions answer is clearly right
- The implementations of the new `Date` object are similar; the prototype for AR Solutions is
  better
- The AR Solutions `DataFrame` omits column names (names), where mine does not; using a list for the
  data in `DataFrame` is a better choice, but harder to implement

```{r}
attributes(data.frame())
```

---

## 15.4.5 Exercises

1.  Add `age()` accessors for the `Person` class.

Answer: the code below adds age accessors to Advanced R `Person`.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Person",
  slots = c(
    name = "character",
    age = "numeric"
  ),
  prototype = list(
    name = NA_character_,
    age = NA_real_
  )
)

john <- new("Person", name = "John Smith")

setGeneric("age", function(x) standardGeneric("age")) |>
  invisible()
setMethod("age", "Person", function(x) x@age)

age(john)

setGeneric("age<-", function(x, value) standardGeneric("age<-")) |>
  invisible()
setMethod("age<-", "Person", function(x, value) {
  x@age <- value
  validObject(x)
  x
})

age(john) <- 51
age(john)
```

AR Solutions: We implement the accessors via an `age()` generic, with a method for the `Person` class and a corresponding replacement function `age<-`:

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setGeneric("age", function(x) standardGeneric("age"))
setMethod("age", "Person", function(x) x@age)

setGeneric("age<-", function(x, value) standardGeneric("age<-"))
setMethod("age<-", "Person", function(x, value) {
  x@age <- value
  validObject(x)
  x
})
```

---

2.  In the definition of the generic, why is it necessary to repeat the
    name of the generic twice?

Answer: according to the R docs, `setGeneric(name)` specifies an existing function. When creating a new generic, the form `setGeneric(name, def)` is used, where `def` is `standardGeneric(name)`.

AR Solutions: Within `setGeneric()` the `name` (1st argument) is needed as the name of the generic. Then, the name also explicitly incorporates method dispatch via `standardGeneric()` within the generic's body (`def` parameter of `setGeneric()`). This behaviour is similar to `UseMethod()` in S3.

---

3.  Why does the `show()` method defined in Section 15.4.3 use
    `is(object)[[1]]`? (Hint: try printing the employee subclass.)

Answer: `is(object)` returns a vector for a subclass, with the current class first.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Employee",
  contains = "Person",
  slots = c(
    boss = "Person"
  ),
  prototype = list(
    boss = new("Person")
  )
)

setMethod("show", "Person", function(object) {
  cat(is(object)[[1]], "\n",
    "  Name: ", object@name, "\n",
    "  Age:  ", object@age, "\n",
    sep = ""
  )
})

john
jane <- new("Employee", name = "Jane Doe")
is(jane)
jane
```

AR Solutions: `is(object)` returns the class of the object. `is(object)` also contains the superclass, for subclasses like `Employee`. In order to always return the most specific class (the subclass), `show()` returns the first element of `is(object)`.

---

4.  What happens if you define a method with different argument names to
    the generic?

Answer: this should cause method dispatch to fail, but I don't know of a way to test.

AR Solutions: It depends. We first create the object `hadley` of class `Person`:

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
.Person <- setClass(
  "Person",
  slots = c(name = "character", age = "numeric")
)
hadley <- .Person(name = "Hadley")
hadley
```

Now let's see which arguments can be supplied to the `show()` generic.

```{r}
formals("show")
```

Usually, we would use this argument when defining a new method.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setMethod("show", "Person", function(object) {
  cat(object@name, "creates hard exercises")
})

hadley
```

When we supply another name as a first element of our method (e.g. `x` instead of `object`), this element will be matched to the correct `object` argument and we receive a warning. Our method will work, though:

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setMethod("show", "Person", function(x) {
  cat(x@name, "creates hard exercises")
})

hadley
```

If we add more arguments to our method than our generic can handle, we will get an error.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
try(
  setMethod("show", "Person", function(x, y) {
    cat(x@name, "is", x@age, "years old")
  })
)
```

If we do this with arguments added to the correctly written `object` argument, we will receive an informative error message. It states that we could add other argument names for generics, which can take the `...` argument.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
try(
  setMethod("show", "Person", function(object, y) {
    cat(object@name, "is", object@age, "years old")
  })
)
```

---

## 15.5.5 Exercises

1.  Draw the method graph for `f(sweat_smile, kissing_cat)`

Answer: skipped.

AR Solutions: Look at the graph and repeat after me: "I will keep my class structure simple and use multiple inheritance sparingly".

---

2.  Draw the method graph for `f(smiley, wink, kissing_smiling_eyes)`

Answer: skipped.

AR Solutions: We see that the method graph below looks simpler than the one above. Relatively speaking, multiple dispatch seems to introduce less complexity than multiple inheritance. Use it with care, though!

---

3.  Take the last example which shows multiple dispatch over two classes that
    use multiple inheritance. What happens if you define a method for all
    terminal classes? Why does method dispatch not save us much work here?

Answer: since all terminal nodes are equidistant from the classes, the method is ambiguous. This is true at the first level as well. The only way to avoid ambiguity is to avoid method dispatch altogether and define a method at the root node.

AR Solutions: We will introduce ambiguity, since one class has distance 2 to all terminal nodes and the other four have distance 1 to two terminal nodes each. To resolve this ambiguity we have to define five more methods, one per class combination.

---

## 15.6.3 Exercises

1.  What would a full `setOldClass()` definition look like for an ordered 
    factor (i.e. add `slots` and `prototype` the definition above)?

Answer: building on the AR Solutions example and the `factor` example in 15.6.1, the following S4 object behaves like the `ordered()` version:

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Ordered",
  contains = "integer",
  slots = c(
    levels = "character",
    ordered = "logical"
  ),
  prototype = structure(
    integer(),
    levels = character(),
    ordered = TRUE
  )
)
setOldClass("ordered", S4Class = "Ordered")

o <- new("Ordered", 1:4, levels = letters[1:4])
o
str(ordered(c("a", "b", "c", "d")))
```

AR Solutions: The purpose of `setOldClass()` lies in registering an S3 class as a "formally defined class", so that it can be used within the S4 object-oriented programming system. When using it, we may provide the argument `S4Class`, which will inherit the slots and their default values (prototype) to the registered class.

Let's build an S4 `OrderedFactor` on top of the S3 factor in such a way.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setOldClass("factor") # use build-in definition for brevity
OrderedFactor <- setClass(
  "OrderedFactor",
  contains = "factor", # inherit from registered S3 class
  slots = c(
    levels = "character",
    ordered = "logical" # add logical order slot
  ),
  prototype = structure(
    integer(),
    levels = character(),
    ordered = logical() # add default value
  )
)
```

We can now register the (S3) ordered-class, while providing an "S4 template". We can also use the S4-class to create new object directly.

```{r, warning = FALSE, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setOldClass("ordered", S4Class = "OrderedFactor")
x <- OrderedFactor(
  c(1L, 2L, 2L),
  levels = c("a", "b", "c"),
  ordered = TRUE
)
str(x)
```

Notes: while my class `Ordered` looks similar to the AR Solutions version, it does not include the S3 class:

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
str(o)
```

---

2.  Define a `length` method for the `Person` class.

Answer: since `Person` is a vector class, its length is the `length()` of any of its fields (we use Person2 here due to prevent conflicts with the already defined Person class):

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
setClass("Person2",
  slots = c(
    name = "character",
    age = "numeric"
  ),
  prototype = list(
    name = NA_character_,
    age = NA_real_
  )
)

setGeneric("length") |>
  invisible()
setMethod("length", "Person2", function(x) length(x@name))

people <- new("Person2", name = c("John Smith", "Jane Doe"), age = c(NA_real_, NA_real_))
people
length(people)
```

AR Solutions: We keep things simple and will just return `"180cm"` when the `length()` method is called on a `Person` object. The method can be defined either as an S3 or S4 method.

```{r, eval = FALSE}
# TODO: eval = FALSE set to allow rendering using build_analysis_site()
length.Person <- function(x) "180cm" # S3
setMethod("length", "Person", function(x) "180cm") # S4
```

---

# 16 Trade-offs

You now know about the three most important OOP toolkits available in R. Now that you understand their basic operation and the principles that underlie them, we can start to compare and contrast the systems in order to understand their strengths and weaknesses. This will help you pick the system that is most likely to solve new problems.

Overall, when picking an OO system, I recommend that you default to S3. S3 is simple, and widely used throughout base R and CRAN. While it's far from perfect, its idiosyncrasies are well understood and there are known approaches to overcome most shortcomings. If you have an existing background in programming you are likely to lean towards R6, because it will feel familiar. I think you should resist this tendency for two reasons. Firstly, if you use R6 it's very easy to create a non-idiomatic API that will feel very odd to native R users, and will have surprising pain points because of the reference semantics. Secondly, if you stick to R6, you'll lose out on learning a new way of thinking about OOP that gives you a new set of tools for solving problems.
