---
title: "Advanced R Workbook (Metaprogramming)"
date: '2023-02-19'
output:
  html_notebook:
    theme:
      version: 5
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
---

Workbook for completing quizzes and exercises from the "Metaprogramming" chapters of [Advanced R](https://adv-r.hadley.nz/index.html), second edition, with comparisons to solutions from [Advanced R Solutions](https://advanced-r-solutions.rbind.io).

```{r setup, message = FALSE, warning = FALSE}
library(lobstr)

# from https://github.com/hadley/adv-r/blob/master/common.R
knitr::opts_chunk$set(
  comment = "#>",
  fig.align = "center"
)

knitr::knit_hooks$set(
  small_mar = function(before, options, envir) {
    if (before) {
      par(mar = c(4.1, 4.1, 0.5, 0.5)) # nolint: undesirable_function_linter.
    }
  }
)
```

# Introduction

This workbook includes answers and solutions to the quizzes and exercises from [Advanced R](https://adv-r.hadley.nz/index.html) and [Advanced R Solutions](https://advanced-r-solutions.rbind.io), organized by chapter. It includes excerpts from both books, copied here.

**WARNING, SPOILERS!** If you haven't read Advanced R and intend to complete the quizzes and exercises, don't read this notebook. It contains my (potentially wrong) answers to both.

# 17 Big picture

Metaprogramming is the hardest topic in this book because it brings together many formerly unrelated topics and forces you grapple with issues that you probably haven't thought about before. You'll also need to learn a lot of new vocabulary, and at first it will seem like every new term is defined by three other terms that you haven't heard of. Even if you're an experienced programmer in another language, your existing skills are unlikely to be much help as few modern popular languages expose the level of metaprogramming that R provides. So don't be surprised if you're frustrated or confused at first; this is a natural part of the process that happens to everyone! 

But I think it's easier to learn metaprogramming now than ever before. Over the last few years, the theory and practice have matured substantially, providing a strong foundation paired with tools that allow you to solve common problems. In this chapter, you'll get the big picture of all the main pieces and how they fit together. 

# 18 Expressions

To compute on the language, we first need to understand its structure. That requires some new vocabulary, some new tools, and some new ways of thinking about R code. The first of these is the distinction between an operation and its result. Take the following code, which multiplies a variable `x` by 10 and saves the result to a new variable called `y`. It doesn't work because we haven't defined a variable called `x`:

```{r, error = TRUE}
rm(x)
try(y <- x * 10)
```

It would be nice if we could capture the intent of the code without executing it. In other words, how can we separate our description of the action from the action itself? 

One way is to use `rlang::expr()`:

```{r}
z <- rlang::expr(y <- x * 10)
z
```

`expr()` returns an expression, an object that captures the structure of the code without evaluating it (i.e. running it). If you have an expression, you can evaluate it with `base::eval()`:

```{r}
x <- 4
eval(z)
y
```

The focus of this chapter is the data structures that underlie expressions. Mastering this knowledge will allow you to inspect and modify captured code, and to generate code with code. We'll come back to `expr()` in Chapter 19, and to `eval()` in Chapter 20.

## 18.2.4 Exercises

1.  Reconstruct the code represented by the trees below:

```{r}
#> █─f
#> └─█─g
#>   └─█─h
#> █─`+`
#> ├─█─`+`
#> │ ├─1
#> │ └─2
#> └─3
#> █─`*`
#> ├─█─`(`
#> │ └─█─`+`
#> │   ├─x
#> │   └─y
#> └─z
```

Answer: code below.

```{r}
lobstr::ast(f(g(h())))
lobstr::ast(1 + 2 + 3)
lobstr::ast((x + y) * z)
```

AR Solutions: Let the source (of the code chunks above) be with you and show you how the ASTs (abstract syntax trees) were produced.

```{r}
ast(f(g(h())))
ast(1 + 2 + 3)
ast((x + y) * z)
```

---

2.  Draw the following trees by hand and then check your answers with
    `lobstr::ast()`.

```{r, eval = FALSE}
f(g(h(i(1, 2, 3))))
f(1, g(2, h(3, i())))
f(g(1, 2), h(3, i(4, 5)))
```

Answer: code below, expression 1:

```{r}
#> █─f
#> └─█─g
#>   └─█─h
#>     └─█─i
#>       └─1
#>       └─2
#>       └─3
lobstr::ast(f(g(h(i(1, 2, 3)))))
```

Expression 2:

```{r}
#> █─f
#> └─1
#> └─█─g
#>   └─2
#>   └─█─h
#>     └─3
#>     └─█─i
lobstr::ast(f(1, g(2, h(3, i()))))
```

Expression 3:

```{r}
#> █─f
#> └─█─g
#> | └─1
#> | └─2
#> └─█─h
#>   └─3
#>   └─█─i
#>     └─4
#>     └─5
lobstr::ast(f(g(1, 2), h(3, i(4, 5))))
```

AR Solutions: Let us delegate the drawing to the `lobstr` package.

---

3.  What's happening with the ASTs below? (Hint: carefully read `?"^"`.)

```{r}
lobstr::ast(`x` + `y`)
lobstr::ast(x ** y) # styler: off
lobstr::ast(1 -> x) # nolint: assignment_linter.
```

Answer: all expressions are changed to their normal forms. In the first expression, the back ticks denoting symbols are not needed. In the third expression, the assignment operator is changed to the standard form. In the second expression, the operator is changed per this note in the R documentation:

> `**` is translated in the parser to `^`, but this was undocumented for many years. It appears as an index entry in Becker *et al* (1988), pointing to the help for `Deprecated` but is not actually mentioned on that page. Even though it had been deprecated in S for 20 years, it was still accepted in R in 2008.

AR Solutions: ASTs start function calls with the name of the function. This is why the call in the first expression is translated into its prefix form. In the second case, `**` is translated by R's parser into `^`. In the last AST, the expression is flipped when R parses it.

---

4.  What is special about the AST below? (Hint: re-read Section
    6.2.1.)

```{r}
lobstr::ast(function(x = 1, y = 2) {}) # nolint: brace_linter.
```

Answer: the AST includes two of the three parts of the function, the `formals()`, `x` and `y`, the `body()`, represented by `{` and `<inline srcref>`.

AR Solutions: The last leaf of the AST is not explicitly specified in the expression. Instead, the `srcref` attribute, which points to the functions source code, is automatically created by base R.

---

5.  What does the call tree of an `if` statement with multiple `else if`
    conditions look like? Why?

Answer: the call tree shows nested if statements; this is because if-else is a function with three arguments, the `cond`, the `cons.expr` and the `alt.expr`. The `alt.expr` for each if-else statement is a new call to if.

```{r}
lobstr::ast(
  if (v == 1) {
    print("one")
  } else if (v == 2) {
    print("two")
  } else if (v == 3) {
    print("three")
  } else if (v == 4) {
    print("four")
  }
)
```

AR Solutions: The AST of nested `else if` statements might look a bit confusing because it contains multiple curly braces. However, we can see that in the `else` part of the AST just another expression is being evaluated, which happens to be an `if` statement and so forth.

```{r}
ast(
  if (FALSE) {
    1
  } else if (FALSE) {
    2
  } else if (TRUE) {
    3
  }
)
```

We can see the structure more clearly if we avoid the curly braces:

```{r}
ast(
  # styler: off
  if (FALSE) 1
  else if (FALSE) 2
  else if (TRUE) 3
  # styler: on
)
```

---
