---
title: "Constraints vs Performance"
author: "John Benninghoff"
date: '2024-04-30'
date-modified: '2024-04-30'
categories: notes
order: 107
output:
  html_notebook:
    theme:
      version: 5
      preset: bootstrap
    css: assets/extra.css
    pandoc_args: --shift-heading-level-by=1
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

Visualizations exploring the use of constraints vs performance improvements in risk management.

# Questions/TODO

- [ ] Questions/TODO list here

```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(jbplot)

xmin <- -5
xmax <- 5
```

# Normal Performance

Replicate a version of Figure 9 from the
[Safety-II White Paper](https://www.england.nhs.uk/signuptosafety/wp-content/uploads/sites/16/2015/10/safety-1-safety-2-whte-papr.pdf),
with help from <https://ggplot2tutor.com/tutorials/sampling_distributions>:

```{r background}
background <- ggplot(data.frame(x = c(xmin, xmax)), aes(x)) +
  scale_x_continuous(breaks = -3:3, minor_breaks = NULL) +
  labs(x = NULL, y = NULL) +
  theme_quo(minor.y = FALSE)

baseline <- stat_function(fun = dnorm, geom = "line")
bad <- stat_function(fun = dnorm, geom = "area", fill = "red", xlim = c(xmin, -2))

background + bad + baseline
```

The plot above shows "bad" outcomes in red. Let's add in "good" outcomes (>1) in green:

```{r good}
good <- stat_function(fun = dnorm, geom = "area", fill = "green", xlim = c(1, xmax))

background + bad + good + baseline
```

# Constrained Performance

One way of reducing "bad" outcomes is by constraining performance - reducing the standard
deviation.

```{r constrained}
constrained <- stat_function(fun = dnorm, args = list(sd = 0.7), geom = "line", color = "blue")
taller <- scale_y_continuous(limits = c(0, 0.6))

background +
  stat_function(
    fun = dnorm, args = list(sd = 0.7), geom = "area", fill = "red", xlim = c(xmin, -2)
  ) +
  stat_function(
    fun = dnorm, args = list(sd = 0.7), geom = "area", fill = "green", xlim = c(1, xmax)
  ) +
  constrained +
  taller
```

Plotting both on the same grid shows the reduction in both "bad" and "good" outcomes:

```{r overlay}
background +
  bad +
  good +
  baseline +
  constrained +
  taller
```

# Improved Performance

Another way of reducing bad outcomes is by improving performance - shifting the mean.

```{r performance}
performance <- stat_function(fun = dnorm, args = list(mean = 1), geom = "line", color = "blue")
improved <- stat_function(
  fun = dnorm, args = list(mean = 1), geom = "area", fill = "green", xlim = c(1, xmax)
)

background +
  stat_function(
    fun = dnorm, args = list(mean = 1), geom = "area", fill = "red", xlim = c(xmin, -2)
  ) +
  improved +
  performance
```

Plotting both together shows a reduction in "bad" and an increase in "good" outcomes:

```{r overlay_perf}
background +
  bad +
  improved +
  baseline +
  performance
```

Comparing all three:

```{r overlay_all}
background +
  bad +
  improved +
  baseline +
  constrained +
  performance +
  taller
```
