---
title: "FaultTree.widget Test"
output:
  html_notebook:
    highlight: textmate
    toc: TRUE
    toc_float:
      collapsed: FALSE
      smooth_scroll: FALSE
  html_document:
    highlight: textmate
---

# Example

The fault tree below is centered in-frame when viewed in RStudio, but is partially off-frame when viewed in a browser, and completely off-frame if the browser window is wide enough. This is true for both `html_notebook` and `html_document`.

```{r tree1}
library(FaultTree)
library(FaultTree.widget)

tree1 <- ftree.make(type="priority",reversible_cond=TRUE, name="Site power loss")
tree1 <- addLogic(tree1, at=1, type="or", name="neither emergency", name2="generator operable")
tree1 <- addLogic(tree1, at=2, type="and", name="Independent failure", name2="of generators")
tree1 <- addLatent(tree1, at=3, mttf=5,mttr=12/8760,inspect=1/26, name="e-gen set fails")
tree1 <- addLatent(tree1, at=3, mttf=5,mttr=12/8760,inspect=1/26, name="e-gen set fails")
tree1 <- addLogic(tree1, at=2, type="inhibit", name="Common cause", name2="failure of generators")
tree1 <- addProbability(tree1, at=6, prob=.05, name="Common cause", name2="beta factor")
tree1 <- addLatent(tree1, at=6, mttf=5,mttr=12/8760,inspect=1/26, name="e-gen set fails")
tree1 <- addDemand(tree1, at=1, mttf=1.0, name="External power", name2="interruption")
 
tree1 <- ftree.calc(tree1)
 
ftree2widget(tree1, height = 500, width = 800)
```

Check rendering without height and width specified:

```{r no_hw}
ftree2widget(tree1)
```

# Solution

A partial fix is to change the line in `inst/htmlwidgets/ftree_widget1.js` and `inst/htmlwidgets/ftree_widget2.js`

from (which sets the position based on the width of the browser window):

```javascript
var width_initial = $(window).width()/2-60;
```

to (which sets the position based on the width of the div containing the widget):

```javascript
var width_initial = $("div").width()/2-60;
```

Then the fault tree will position itself within view, but not always centered. A complete fix would render the fault tree centered within the widget.

Other approaches tested:

```javascript
// var width_initial = $('div:contains("htmlwidget")').width()/2-60;
// var width_initial = $("div").filter(function(){ return this.children.length === 0; }).width()/2-60;
```
