---
title: "ggplot2 (Getting started)"
date: '2022-11-20'
output:
  html_notebook:
    theme:
      version: 5
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
---

Workbook for completing quizzes and exercises from the "Getting started" chapters of [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html), third edition, with comparisons to solutions from [Solutions to ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book-solutions-3ed.netlify.app).

```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(dplyr)

knitr::opts_chunk$set(
  comment = "#>",
  fig.align = "center"
)
```

# Introduction

This workbook includes answers and solutions to the quizzes and exercises from [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html) and [Solutions to ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book-solutions-3ed.netlify.app), organized by chapter. It includes excerpts from both books, copied here.

**WARNING, SPOILERS!** If you haven't read the ggplot2 book and intend to complete the quizzes and exercises, don't read this notebook. It contains my (potentially wrong) answers to both.

# 2 First steps

The goal of this chapter is to teach you how to produce useful graphics with ggplot2 as quickly as possible. You’ll learn the basics of ggplot() along with some useful “recipes” to make the most important plots. ggplot() allows you to make complex plots with just a few lines of code because it’s based on a rich underlying theory, the grammar of graphics. Here we’ll skip the theory and focus on the practice, and in later chapters you’ll learn how to use the full expressive power of the grammar.

## 2.2.1 Exercises

1.  List five functions that you could use to get more information about the
    `mpg` dataset.

Answer:

Five common base R functions I use to interrogate dataframes are:

1. `summary()`
1. `hist()`
1. `plot()`
1. `lm()`
1. `View()`

GG Solutions:

-   `help(mpg)`: Documentation of dataset
-   `dim(mpg)`: Dimensions of dataset
-   `summary(mpg)`: Summary measures of dataset
-   `str(mpg)`: Display of the internal structure of dataset
-   `glimpse(mpg)`: `dplyr` version of `str(mpg)`

---

2.  How can you find out what other datasets are included with ggplot2?

Answer: use `data(package = "ggplot2")`, which returns the following:

```
Data sets in package ‘ggplot2’:

diamonds                      Prices of over 50,000 round cut diamonds
economics                     US economic time series
economics_long                US economic time series
faithfuld                     2d density estimate of Old Faithful data
luv_colours                   'colors()' in Luv space
midwest                       Midwest demographics
mpg                           Fuel economy data from 1999 to 2008 for 38 popular models of
                              cars
msleep                        An updated and expanded version of the mammals sleep dataset
presidential                  Terms of 12 presidents from Eisenhower to Trump
seals                         Vector field of seal movements
txhousing                     Housing sales in TX
```

GG Solutions:

`data(package = "ggplot2")` loads the available data sets in ggplot2. Alternatively,if you have internet access, go to https://ggplot2.tidyverse.org/reference/index.html#section-data

---
    
3.  Apart from the US, most countries use fuel consumption (fuel consumed
    over fixed distance) rather than fuel economy (distance travelled with
    fixed amount of fuel). How could you convert `cty` and `hwy` into the
    European standard of l/100km? 

Answer: to convert miles/gallon to liters/100km, we use the formula:

$\frac{1}{miles/gallon} \times \frac{3.78541}{1.609344 \div 100} = l/100km$

```{r}
mpg_l100km <- function(x) {
  (1 / x) * 3.78541 / (1.609344 / 100)
}

mpg |>
  mutate(cty = mpg_l100km(cty), hwy = mpg_l100km(hwy))
```

GG Solutions:

- According to [asknumbers](https://www.asknumbers.com/mpg-to-L100km.aspx), you divide 235.214583 by the mpg values in `cty` and `hwy` to convert them into the European standard of l/100km. 

- Function to convert into European standard (Rademaker, 2016):

```{r}
# nolint start: object_name_linter.
mpgTol100km <- function(milespergallon) {
  GalloLiter <- 3.785411784
  MileKilometer <- 1.609344

  l100km <- (100 * GalloLiter) / (milespergallon * MileKilometer)
  l100km
}
# nolint end
```

---
    
4.  Which manufacturer has the most models in this dataset? Which model has 
    the most variations? Does your answer change if you remove the redundant
    specification of drive train (e.g. "pathfinder 4wd", "a4 quattro") from the 
    model name?

Answer: we can generate a dataframe with manufacturers, models, and variations using:

```{r}
mpg_summary <- mpg |>
  count(manufacturer, model) |>
  rename(variations = n)

mpg_summary
```

The model with the most variations is the Dodge Caravan 2wd.

```{r}
mpg_summary |>
  arrange(variations) |>
  tail(1)
```
The manufacturer with the most models is Toyota.

```{r}
mpg_summary |>
  count(manufacturer) |>
  arrange(n) |>
  tail(1)
```

If we first remove the redundant drive train from the summary, we get:

```{r}
mpg_summary2 <- mpg |>
  mutate(model = gsub(" (quattro)|(2wd)|(4wd)|(awd)$", "", model)) |>
  count(manufacturer, model) |>
  rename(variations = n)

mpg_summary2
```

The model with the most variations has changed to the Audi A4.

```{r}
mpg_summary2 |>
  arrange(variations) |>
  tail(1)
```

The manufacturer with the most models is still Toyota.

```{r}
mpg_summary2 |>
  count(manufacturer) |>
  arrange(n) |>
  tail(1)
```

GG Solutions:

Not included here because I believe the answer to question 4 in [2.2 Exercises](https://ggplot2-book-solutions-3ed.netlify.app/first-steps.html#exercises-1) is incorrect, as it counts variants by manufacturer, not models by manufacturer.

---

## 2.3.1 Exercises

1.  How would you describe the relationship between `cty` and `hwy`?
    Do you have any concerns about drawing conclusions from that plot?

Answer: we can explore the relationship using a scatterplot: 

```{r}
ggplot(mpg, aes(cty, hwy)) +
  geom_point()
```

There is a strong correlation between `cty` and `hwy`, which is expected since these are both measures of fuel efficiency of the same car.

GG Solutions:

- The plot shows a strongly linear relationship, which tells me that `cty` and `hwy` are highly correlated variables. The only concern I have is that the points seem to be overlapping.
- There is not much insight to be gained except that cars which are fuel efficient on a highway are also fuel efficient in cities. This relationship is probably a function of speed (Rademaker, 2016) 

--

2.  What does `ggplot(mpg, aes(model, manufacturer)) + geom_point()` show?
    Is it useful? How could you modify the data to make it more informative?

Answer: this plot shows each model as a point for each manufacturer, which is not particularly useful, although it does allow for a quick count of models by manufacturer. A bar chart showing count of models by manufacturer would be more useful.

```{r}
ggplot(mpg, aes(model, manufacturer)) +
  geom_point()
```

GG Solutions:

- The plot shows the manufacturer of each model. Its not very readable since there are too many models and this clutters up the x-axis with too many ticks! I would just plot 20 or so models so that the graph is more readable. See below:

```{r}
mpg %>%
  head(25) %>%
  ggplot(aes(model, manufacturer)) +
  geom_point()
```

- A possible alternative would be to look total number of observations for each manufacturer-model combination using geom_bar(). (Rademaker, 2016)

```{r}
df <- mpg %>%
  transmute("man_mod" = paste(manufacturer, model, sep = " "))
ggplot(df, aes(man_mod)) +
  geom_bar() +
  coord_flip()
```

Note: I prefer my solution for this exercise.

--

3.  Describe the data, aesthetic mappings and layers used for each of the 
    following plots. You'll need to guess a little because you haven't seen
    all the datasets and functions yet, but use your common sense! See if you 
    can predict what the plot will look like before running the code.
    
    1. `ggplot(mpg, aes(cty, hwy)) + geom_point()`
    1. `ggplot(diamonds, aes(carat, price)) + geom_point()`
    1. `ggplot(economics, aes(date, unemploy)) + geom_line()`
    1. `ggplot(mpg, aes(cty)) + geom_histogram()`

Answer:

1. `ggplot(mpg, aes(cty, hwy)) + geom_point()` - scatterplot of city mpg against highway mpg
1. `ggplot(diamonds, aes(carat, price)) + geom_point()` - scatterplot of diamond weight in carats
   against price
1. `ggplot(economics, aes(date, unemploy)) + geom_line()` - a line plot showing the unemployment
   rate over time
1. `ggplot(mpg, aes(cty)) + geom_histogram()` - a histogram of city mpg

```{r}
ggplot(mpg, aes(cty, hwy)) +
  geom_point()
ggplot(diamonds, aes(carat, price)) +
  geom_point()
ggplot(economics, aes(date, unemploy)) +
  geom_line()
ggplot(mpg, aes(cty)) +
  geom_histogram()
```

In general, my guesses are correct (given my existing experience with ggplot), except for the economics plot, which appears to show the number of unemployed people instead of an employment rate.

GG Solutions: (Rademaker, 2016)

1. `ggplot(mpg, aes(cty, hwy)) + geom_point()`
-   *Data*: `mpg`
-   *Aesthetic*: highway miles per gallon is mapped to y position and city miles per gallon is mapped to x position.
-   *Layer*: points

2. `ggplot(diamonds, aes(carat, price)) + geom_point()`
-   *Data*: `diamonds`
-   *Aesthetic*: price in US dollars is mapped to y position, weight of the diamond is mapped to x position.
-   *Layer*: points

3.  `ggplot(economics, aes(date, unemploy)) + geom_line()`
-   *Data*: `economics`
-   *Aesthetic*: median duration of unemployment, in weeks, is mapped to y position and month of data collection is mapped to x position.
-   *Layer*: line

---

## 2.4.1 Exercises

1.  Experiment with the colour, shape and size aesthetics. What happens when 
    you map them to continuous values? What about categorical values? What 
    happens when you use more than one aesthetic in a plot?

Answer: observations based on the plots below:

- Using a continuous variable with shape generates an error
- Using a categorical variable with shape for more than 6 values generates a warning
- Using a continuous variable with size works well
- Using a categorical variable with size generates a warning
- Using a continuous variable with color creates a gradient
- Using a categorical variable with color creates a discrete mapping
- Multiple aesthetics can overload the plot

```{r}
# continuous
ggplot(mpg, aes(displ, hwy, color = cyl)) +
  geom_point()
# categorical
ggplot(mpg, aes(displ, hwy, color = drv)) +
  geom_point()
# multiple
mpg |>
  mutate(year = as.factor(year)) |>
  ggplot(aes(displ, hwy, color = drv, size = cyl, shape = year)) +
  geom_point()
```

GG Solutions offers a similar and less complete answer.

---

2.  What happens if you map a continuous variable to shape? Why? What happens
    if you map `trans` to shape? Why?

Answer: As noted above, using a continuous variable with shape generates an error, because plots must have discrete shapes. Using a categorical variable with shape for more than 6 values (like `trans`) generates a warning, since (as the error notes) many different shapes can be hard to distinguish on a plot.

GG Solutions presents a similar answer.

---

3.  How is drive train related to fuel economy? How is drive train related to
    engine size and class?

Answer: The plots below show that:

- Front wheel drive is correlated with higher fuel economy
- Front wheel drive is correlated with smaller displacement, rear wheel drive is correlated with
  larger displacement, and four wheel drive is not correlated with displacement
- Some classes only have a single drive train:
  - Two seaters only have rear wheel drive
  - Minivans only have front wheel drive
  - Pickups only have four wheel drive

```{r}
ggplot(mpg, aes(hwy, drv)) +
  geom_point()
ggplot(mpg, aes(displ, drv)) +
  geom_point()
ggplot(mpg, aes(class, drv)) +
  geom_point()
```

GG Solutions presents a similar answer using bar charts, and compares engine size and class together in a single plot.

Note: GG Solutions does not explore the relationship between drive train and class.

---
