---
title: "Hockey Cards Analysis"
author: "John Benninghoff"
date: '2024-04-17'
date-modified: '2024-04-17'
categories: notes
order: 106
output:
  html_notebook:
    theme:
      version: 5
      preset: bootstrap
    css: assets/extra.css
    pandoc_args: --shift-heading-level-by=1
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

A simple Monte Carlo simulation in R, replicating Julia code from a LinkedIn [post](https://www.linkedin.com/posts/etorkia_gretzky-card-simulation-using-julia-activity-7186379288563048448--FVH).

```{r setup, message = FALSE, warning = FALSE}
library(ggplot2)
library(jbplot)
```

# Background

I came across an interesting
[post](https://www.linkedin.com/posts/etorkia_gretzky-card-simulation-using-julia-activity-7186379288563048448--FVH)
on LinkedIn that used Monte Carlo simulation to help answer the question "How much is a box of
unopened Canadian hockey cards worth?" The example code was in Julia, and I wanted to recreate it in
R for comparison.

# Code

The base R code below is functionally equivalent to the Julia code, except it omits the trial ID in
the result:

```{r gretsky_cards}
gretsky_cards <- function(trials) {
  set_cards <- 396
  carton_cards <- 672
  cartons <- 16
  box_cards <- carton_cards * cartons

  replicate(trials, {
    carton <- sample(1:set_cards, box_cards, replace = TRUE)
    sum(carton == 99)
  })
}

gretskys <- gretsky_cards(10000)
```

Also, instead of using a loop, I used `replicate()`, which I think is easier to use and understand.

# Answer - Base R

Replicating the answer in base R:

```{r summary}
summary(gretskys)
hist(gretskys)
```

# Answer - ggplot2

Use ggplot2 to create a prettier histogram:

```{r histogram}
ggplot(as.data.frame(gretskys), aes(gretskys)) +
  geom_hist_bw(binwidth = 1) +
  labs(title = "Gretsky cards per case", x = NULL, y = NULL) +
  labs(caption = "Number of occurrences over 10,000 simulations") +
  theme_quo()
```

# Performance

How does the performance compare to Julia?

```{r benchmark}
bench::mark(gretsky_cards(10000))
```

In this case, it certainly appears to be slower than the Julia test result which was 1.63 seconds.
