---
title: "R Books"
author: "John Benninghoff"
date: '2023-11-17'
file-modified: '2023-11-17'
categories: reading
order: 1
output:
  html_notebook:
    theme:
      version: 5
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

An actively maintained and curated list of R Books and other recommended resources from my reading list.

## Libraries

Libraries used in this notebook.

```{r setup, message = FALSE, warning = FALSE}
library(palmerpenguins)
library(ggplot2)
library(ggthemes)
```

## Getting Started

Books and resources I recommend for learning R.

### R for Data Science

[R for Data Science (2e)](https://r4ds.hadley.nz) - Second Edition of
[Hadley Wickham's](https://hadley.nz/) introductory book on R and exploratory data analysis. The
book contains example code and exercises in most chapters. I used the
[first edition](https://r4ds.had.co.nz/) as my primary source when I was first learning R, and
still recommend the book for beginners who want to learn R. The book teaches the basics of using
tidyverse R^[For a detailed explanation of "tidyverse R", see [R Dialects]] for exploratory data
analysis and visualization. A companion book provides suggested
[solutions](https://mine-cetinkaya-rundel.github.io/r4ds-solutions) to the exercises.

Sample visualization from the first chapter:

```{r r4ds, message = FALSE, warning = FALSE}
# visualization from https://r4ds.hadley.nz/data-visualize#first-steps
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) +
  scale_color_colorblind()
```

### Resources

In addition to R for Data Science, I recommend the following resources for new R users:

- [The R Graph Gallery](https://r-graph-gallery.com/) - library of charts made with R and ggplot2
- Packages for writing better code:
  - [renv](https://rstudio.github.io/renv/) - for package version management by project
  - [styler](https://styler.r-lib.org/) - for consistent formatting of R code
  - [lintr](https://lintr.r-lib.org) - for avoiding common mistakes
- Style guides for writing better code:
  - [tidyverse style guide](https://style.tidyverse.org/) - implemented by styler
  - [Google's R Style Guide](https://google.github.io/styleguide/Rguide.html) - a fork of the
    tidyverse guide

[rdev](https://jabenninghoff.github.io/rdev/index.html), my personalized collection of R development
tools, includes all three of these packages and more, along with my own
[style guide](https://jabenninghoff.github.io/rdev/articles/style-guide.html) and R environment
setup [instructions](https://jabenninghoff.github.io/rdev/articles/rdev.html).

## Additional Reading

Books I read to improve my knowledge of R.

- [R Packages (2e)](https://r-pkgs.org/) - the second edition of Hadley Wickham's book on writing R
  packages, which I used to build [rdev](https://github.com/jabenninghoff/rdev)
- [Advanced R](https://adv-r.hadley.nz/) - the second edition of Hadley's book on R programming,
  which explains the R language (the [first edition](http://adv-r.had.co.nz) uses more base R than
  the second edition)
- [Advanced R Solutions](https://advanced-r-solutions.rbind.io) - solutions to exercises from
  *Advanced R*
- [ggplot2: Elegant Graphics for Data Analysis (3e)](https://ggplot2-book.org) - the third edition
  of Hadley's book on his implementation of Leland Wilkinson's *Grammar of Graphics*
- [Solutions to ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book-solutions-3ed.netlify.app) -
  solutions to exercises from *ggplot2*

My own notes and solutions to the Advanced R and ggplot2 exercises are available on this site.

### raps-with-r

[Building reproducible analytical pipelines with R](https://raps-with-r.dev) - the stated goal of
the book is to improve the reproducibility of data analysis. I don't recommend this book. Section 1
is a reasonably good introduction to git and GitHub, but introduces trunk-based development without
covering linear commit history. Section 2 provides some good advice, but much that I disagree with,
including inline use of `knitr::knit_child` to automate creation of sections (which break the
notebook workflow) and `fusen` to create packages from `.Rmd` files, which I found to create extra
work with no clear benefits over using and/or extending the standard package layout like
[vertical](https://www.crumplab.com/vertical/index.html) or my own
[layout](https://jabenninghoff.github.io/rdev/articles/analysis-package-layout.html) from
[rdev](https://jabenninghoff.github.io/rdev/index.html). (Interestingly, the author of vertical is
[also](https://www.crumplab.com/blog/post_886_10_14_22_quartoProjects/)
[switching](https://www.crumplab.com/blog/663_quarto_projects/) to [Quarto](https://quarto.org) for
reproducible research and analysis) I also found the code examples to be inconsistent and a bit
clunky.

<!-- TODO: add notes on chapters 12-15 -->

I do follow the book's core recommendations for reproducibility:

- Use version control extensively
- Use trunk-based development
- Adopt functional programming
- Use R Markdown which embeds analysis code and code output directly into the written report
- Use [renv](https://rstudio.github.io/renv/) to manage dependencies
- Package all R code and publish using GitHub Pages

The book does reference some interesting reading I've added to my list:

- John M. Chambers. "Object-Oriented Programming, Functional Programming and R."
  Statist. Sci. 29 (2) 167 - 180, May 2014. <https://doi.org/10.1214/13-STS452>
- [Trunk-Based Development And Branch By Abstraction](https://trunkbaseddevelopment.com/)

While it isn't referenced in the book, I discovered [vertical](https://www.crumplab.com/vertical/)
while reading it, and will be reviewing and comparing it to my own
[rdev](https://jabenninghoff.github.io/rdev/).

## Future Reading

R books on my reading list.

- [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/) - written by the author
  of [knitr](https://yihui.org/knitr/)
- [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/) - the follow-up to *The
  Definitive Guide*

## R Dialects

An explanatory note on the R dialects of base R and tidyverse R.

The [R programming language](https://en.wikipedia.org/wiki/R_(programming_language)) is over 30
years old and has a large number of [packages](https://en.wikipedia.org/wiki/R_package) (R
libraries) that extend R. Unlike
[python](https://en.wikipedia.org/wiki/Python_(programming_language)) (a general purpose language),
R was designed specifically for analysis, visualization, and statistical modeling, which is why I
[chose R](https://www.information-safety.org/2020/09/11/working-with-r/) for data analysis: it has
built-in support for data structures like data frames (implemented in python using
[pandas](https://en.wikipedia.org/wiki/Pandas_(software))), vectors, packages for just about any
statistical tool you'd need, and of course, [ggplot2](https://en.wikipedia.org/wiki/Ggplot2). In
fairness, python is more popular, more robust, and a better tool for some tasks, like data
acquisition and machine learning (which were not priorities for my use).

Like many human languages, R has developed two distinct dialects: base R and tidyverse R. Base R
consists of the [packages](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-in-R)
included in the R distribution
*(`r paste(rownames(utils::installed.packages(priority = "base")), collapse = ", ")`)*, and the
[Tidyverse](https://en.wikipedia.org/wiki/Tidyverse) is a collection of packages that implement a
[domain-specific language](https://en.wikipedia.org/wiki/Domain-specific_language) for data
analysis, originally created by [Hadley Wickham](https://en.wikipedia.org/wiki/Hadley_Wickham).

In my experience, tidyverse R is better for data analysis, where base R is better for writing
packages - tidyverse functions are closer to natural language, but have many more dependencies.
[Comparing](https://atrebas.github.io/post/2019-03-03-datatable-dplyr/) two popular tools for data
manipulation, [dplyr](https://dplyr.tidyverse.org) (tidyverse R) and
[data.table](https://rdatatable.gitlab.io/data.table/) (base R) shows these differences.

This code snippet is from a short analysis of survey responses using dplyr:

```r
survey_results <- survey_import |>
  mutate(across(Q1:Q7, ~ case_when(
    .x == "strongly disagree" ~ 1,
    .x == "disagree" ~ 2,
    .x == "neither agree nor disagree" ~ 3,
    .x == "agree" ~ 4,
    .x == "strongly agree" ~ 5
  ))) |>
  mutate(Q8 = as.numeric(Q8 == "Yes")) |>
  arrange(end_date)
```

The code is reasonably easy to understand, even if you're not familiar with R.

The same code written in data.table isn't as clear:

```r
likert_5 <- c(
  "strongly disagree", "disagree", "neither agree nor disagree",
  "agree", "strongly agree"
)
q_likert <- paste0("Q", 1:7)
q_yesno <- "Q8"

survey_results <- copy(survey_import)
survey_results <- survey_results[
  , (q_likert) := lapply(.SD, \(x) as.numeric(factor(x, levels = likert_5))),
  .SDcols = q_likert
][
  , (q_yesno) := lapply(.SD, \(x) as.numeric(x == "Yes")),
  .SDcols = q_yesno
][
  order(end_date)
]
```

While it may be harder to read, data.table has some clear advantages: it is quite
[fast](https://h2oai.github.io/db-benchmark/), especially with very large datasets, and has no
[dependencies](https://github.com/Rdatatable/data.table/blob/master/DESCRIPTION) other than base R,
where dplyr has many.

These tradeoffs are why I tend to use tidyverse R for analysis and base R for functions (most
tidyverse expressions have functional equivalents in base R). Code used in data analysis should be
clear and easy to read, which tidyverse R excels at. Packaged functions provide documentation and
the source code isn't typically read, but many dependencies can be problematic; `R CMD check` will
raise a NOTE if there are too many imports.
