---
title: "R Books"
author: "John Benninghoff"
date: '2023-11-17'
date-modified: '2025-11-12'
categories: reading
order: 1
output:
  html_notebook:
    theme:
      version: 5
      preset: bootstrap
    css: assets/extra.css
    pandoc_args: --shift-heading-level-by=1
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
---

An actively maintained and curated list of R Books and other recommended resources from my reading list.

# Libraries

Libraries used in this notebook.

```{r setup, message = FALSE, warning = FALSE}
library(palmerpenguins)
library(ggplot2)
library(ggthemes)
library(showtext)
library(cranlogs)
library(dplyr)
```

# Getting Started

Books and resources I recommend for learning R.

## R for Data Science

[R for Data Science (2e)](https://r4ds.hadley.nz) - Second Edition of
[Hadley Wickham's](https://hadley.nz/) introductory book on R and exploratory data analysis. The
book contains example code and exercises in most chapters. I used the
[first edition](https://r4ds.had.co.nz/) as my primary source when I was first learning R, and
still recommend the book for beginners who want to learn R. The book teaches the basics of using
tidyverse R^[For a detailed explanation of "tidyverse R", see [R Dialects]] for exploratory data
analysis and visualization. A companion book provides suggested
[solutions](https://mine-cetinkaya-rundel.github.io/r4ds-solutions/) to the exercises.

Sample visualization from the first chapter:

```{r r4ds, message = FALSE, warning = FALSE}
# visualization from https://r4ds.hadley.nz/data-visualize#first-steps
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  ) +
  scale_color_colorblind()
```

## Resources

In addition to R for Data Science, I recommend the following resources for new R users:

- [The R Graph Gallery](https://r-graph-gallery.com/) - library of charts made with R and ggplot2
- Packages for writing better code:
  - [renv](https://rstudio.github.io/renv/) - for package version management by project
  - [styler](https://styler.r-lib.org/) - for consistent formatting of R code
  - [lintr](https://lintr.r-lib.org) - for avoiding common mistakes
- Style guides for writing better code:
  - [tidyverse style guide](https://style.tidyverse.org/) - implemented by styler
  - [Google's R Style Guide](https://google.github.io/styleguide/Rguide.html) - a fork of the
    tidyverse guide

[rdev](https://jabenninghoff.github.io/rdev/index.html), my personalized collection of R development
tools, includes all three of these packages and more, along with my own
[style guide](https://jabenninghoff.github.io/rdev/articles/style-guide.html) and R environment
setup [instructions](https://jabenninghoff.github.io/rdev/articles/rdev.html).

## Custom Fonts

Working with custom fonts in R can be problematic, and font support is platform-dependent. By
default, R provides default mappings for three device-independent font families: `sans` (Helvetica or Arial), `serif` (Times New Roman), and `mono` (Courier New). On macOS, `quartz` is the default
graphics device, and uses the following fonts:

```{r quartzFonts}
quartzFonts()
```

Custom fonts installed on the local system *should* work without installing any additional packages:

```{r system_fonts}
size <- 28

ggplot() +
  annotate("text", x = 0, y = 0, label = "Default", size = size) +
  annotate("text", x = 0, y = -1, label = "Lato", family = "Lato", size = size) +
  annotate("text", x = 0, y = -2, label = "Metropolis", family = "Metropolis", size = size) +
  scale_y_continuous(limits = c(-2.5, 0.5)) +
  theme_void()
```

This approach works, but isn't portable as it introduces hidden dependencies on the custom fonts.
The plot above works because I have the [Lato](https://www.latofonts.com/) and
[Metropolis](https://www.1001fonts.com/metropolis-font.html) fonts installed on the macOS system I
use to build this page; on systems without these fonts installed this plot would generate warnings
and fallback to the default font:

```{r missing}
ggplot() +
  annotate("text", x = 0, y = 0, label = "Not Missing", size = size) +
  annotate("text", x = 0, y = -1, label = "Missing", family = "missing", size = size) +
  scale_y_continuous(limits = c(-1.5, 0.5)) +
  theme_void()
```

There are a few packages that facilitate use of custom fonts in R:

- [extrafont](https://cran.r-project.org/package=extrafont) is an older package that can import
  fonts for use in PDF, PostScript, and Windows
- [showtext](https://cran.r-project.org/package=showtext) is a newer package that is most often
  recommended for custom fonts, and can import fonts directly from Google Fonts with
  `font_add_google()`
- [ragg](https://cran.r-project.org/package=ragg) is a graphics back-end that gives access to all
  installed system fonts, and is recommended as an alternative to showtext
- [systemfonts](https://cran.r-project.org/package=systemfonts) locates installed fonts using system
  libraries on macOS and Linux, and Freetype on Windows

On macOS, no additional packages are needed to access all system fonts, although this may not be the
case on other platforms (I haven't tested). I currently use a combination of base R and showtext.
systemfonts can help troubleshoot issues by showing what system fonts R recognizes.

showtext will work even with fonts not present on the local system:

```{r showtext}
font_add_google("Open Sans")
showtext_auto()

ggplot() +
  annotate("text", x = 0, y = 0, label = "Default", size = size) +
  annotate("text", x = 0, y = -1, label = "Open Sans", family = "Open Sans", size = size) +
  scale_y_continuous(limits = c(-1.5, 0.5)) +
  theme_void()
```

An important caveat: showtext *replaces* all system fonts when enabled. To disable it and use the
standard font system, run:

```{r disable_showtext}
showtext_auto(FALSE)
```

Additional articles on using custom fonts in R:

- <https://r-graph-gallery.com/custom-fonts-in-R-and-ggplot2.html>
- <https://gradientdescending.com/adding-custom-fonts-to-ggplot-in-r/>
- <https://www.cararthompson.com/posts/2024-01-12-using-fonts-in-r-for-dataviz/2024-01-12_getting-fonts-to-work>

# Additional Reading

Books I read to improve my knowledge of R.

- [R Packages (2e)](https://r-pkgs.org/) - the second edition of Hadley Wickham's book on writing R
  packages, which I used to build [rdev](https://github.com/jabenninghoff/rdev)
- [Advanced R](https://adv-r.hadley.nz/) - the second edition of Hadley's book on R programming,
  which explains the R language (the [first edition](http://adv-r.had.co.nz) uses more base R than
  the second edition)
- [Advanced R Solutions](https://advanced-r-solutions.rbind.io) - solutions to exercises from
  *Advanced R*
- [ggplot2: Elegant Graphics for Data Analysis (3e)](https://ggplot2-book.org) - the third edition
  of Hadley's book on his implementation of Leland Wilkinson's *Grammar of Graphics*
- [Solutions to ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book-solutions-3ed.netlify.app) -
  solutions to exercises from *ggplot2*

My own notes and solutions to the Advanced R and ggplot2 exercises are available on this site.

## raps-with-r

[Building reproducible analytical pipelines with R](https://raps-with-r.dev) - the stated goal of
the book is to improve the reproducibility of data analysis. I don't recommend this book. Section 1
is a reasonably good introduction to git and GitHub, but introduces trunk-based development without
covering linear commit history. Section 2 provides some good advice, but much that I disagree with,
including inline use of `knitr::knit_child` to automate creation of sections (which break the
notebook workflow) and `fusen` to create packages from `.Rmd` files, which I found to create extra
work with no clear benefits over using and/or extending the standard package layout like
[vertical](https://www.crumplab.com/vertical/index.html) or my own
[layout](https://jabenninghoff.github.io/rdev/articles/analysis-package-layout.html) from
[rdev](https://jabenninghoff.github.io/rdev/index.html). (Interestingly, the author of vertical is
[also](https://www.crumplab.com/blog/post_886_10_14_22_quartoProjects/)
[switching](https://www.crumplab.com/blog/663_quarto_projects/) to [Quarto](https://quarto.org) for
reproducible research and analysis) I also found the code examples to be inconsistent and a bit
clunky.

Chapter 12 gives recommendations on testing: unit testing with some basic examples,
[assertive programming](https://en.wikipedia.org/wiki/Assertion_(software_development)),
[Test-Driven Development (TDD)](https://en.wikipedia.org/wiki/Test-driven_development), and test
coverage. It suggests three packages for R assertions:
[`assertthat`](https://github.com/hadley/assertthat),
[`chk`](https://poissonconsulting.github.io/chk/), and
[`checkmate`](https://mllg.github.io/checkmate/). Both `chk` and `checkmate` are designed to check
function arguments; of the two, `checkmate` appears to be more robust and built to address the
downside of R being a dynamically typed language.

For data validation, I currently use `stopifnot()`, although I may switch to either `assertr` or
`validate`, which are both popular packages. I validate function arguments using manual checks, but
`checkmate` looks appealing as a way to write more succinct code. Total downloads for the `assertr`,
`validate`, `chk`, and `checkmate` packages for the last month are listed below:

```{r cran_downloads}
cran_downloads(
  packages = c("assertr", "validate", "chk", "checkmate"),
  when = "last-month"
) |>
  group_by(package) |>
  summarize(downloads = sum(count), as_of = max(date))
```

2023-12-29 Update: I've started using `checkmate` to check function arguments and have found
that `validate` is better overall at data validation.

Chapter 13 introduces [targets](https://docs.ropensci.org/targets/), "a Make-like pipeline tool for
statistics and data science in R." Conceptually, targets is very similar to an R Notebook, but like
Make, will skip components that are up to date, and can run targets in parallel to speed up builds.
targets can also render R Markdown documents using the `tarchetypes` package. I found the example
target pipeline in the book convoluted and didn't attempt to follow it. The targets
[manual](https://books.ropensci.org/targets/) explains that it implements
[two kinds](https://books.ropensci.org/targets/literate-programming.html#scope) of literate
programming:

> 1. A literate programming source document (or Quarto project) that renders inside an individual
>    target. Here, you define a special kind of target that runs a lightweight R Markdown report
>    which depends on upstream targets.
> 1. Target Markdown, an overarching system in which one or more Quarto or R Markdown files write
>    the _targets.R file and encapsulate the pipeline.

Of these two types, the second is a better fit with my preferred workflow: including multiple
self-contained notebooks in a single repository. From the
[appendix](https://books.ropensci.org/targets/markdown.html), the design of Target Markdown supports
both interactive development using notebooks and running a pipeline non-interactively when rendering
the final document. For my work, targets doesn't offer significant advantages over using R
Markdown and Quarto - the only slowdown I typically encounter is when building a site with many
notebooks, which the Quarto [freeze](https://quarto.org/docs/projects/code-execution.html#freeze)
option handles by only re-rendering documents when the source file changes. (I'm not performing the
large, complex computations that targets is designed for)

Chapter 14 covers Docker, and asserts that it is necessary for reproducibility. While using Docker
ensures a stable operating system environment, I think the book overstates the case for
reproducibility, citing a single example where the operating system changed the results of an
analysis because the scripts relied on OS file ordering. Much like use of targets, Docker would be
most useful for projects with complex development environments shared across teams, but much of the
benefits can be achieved using other approaches, like using [OS-independent](https://fs.r-lib.org)
code. The chapter also encourages using a "frozen" development environment that is updated on a
fixed schedule to minimize the impact of frequent updates. This is exactly the opposite of the
modern DevOps approach - the issues created by small, frequent updates are smaller and easier to
address than the more complex problems created by large upgrades. I *always* start development by
updating packages to the latest production release - while this sometimes introduces issues, they
are typically easy to fix, and prioritizing maintenance first improves quality and security.

An alternate approach to using Docker is to leverage
[GitHub Actions](https://github.com/features/actions), which provides on-demand virtual machines
that can be used to consistently test, build, and deploy code. The Tidyverse community provides a
[library](https://github.com/r-lib/actions) of GitHub Actions for R, which I've customized for
[rdev](https://jabenninghoff.github.io/rdev/reference/use_rdev_package.html). In fact, chapter 15
covers use of GitHub Actions using r-lib and presents them as a potential alternative for Docker.

Overall, I do follow the book's core recommendations for reproducibility:

- Use version control extensively
- Use trunk-based development
- Adopt functional programming and [DRY](https://en.wikipedia.org/wiki/Don't_repeat_yourself)
- Use R Markdown which embeds analysis code and code output directly into the written report
- Use [renv](https://rstudio.github.io/renv/) to manage dependencies
- Package all R code and publish using GitHub Pages
- Write unit tests using [testthat](https://testthat.r-lib.org)
- Use assertions to validate function arguments and imported data
- Check test coverage using [`covr`](https://covr.r-lib.org)
- Use automated builds (but using Quarto instead of targets)

I would consider use of targets and Docker for larger, more complex, or long-lived projects, but I
found that `fusen` offered no clear benefits and wouldn't recommend its use. I think the biggest
lesson I took from the book was to follow DRY a bit more strictly than I currently do, and use more
functions, tests, and assertions in my analysis code.

The book does reference some interesting reading I've added to my list:

- John M. Chambers. "Object-Oriented Programming, Functional Programming and R."
  Statist. Sci. 29 (2) 167 - 180, May 2014. <https://doi.org/10.1214/13-STS452>
- [Trunk-Based Development And Branch By Abstraction](https://trunkbaseddevelopment.com/)

While reading the book, I discovered some interesting additional resources:

- [Vertical](https://www.crumplab.com/vertical/), a "an R-based structured workflow for creating
  and sharing research assets in the form of an extended R package", which I plan to review and
  compare to [rdev](https://jabenninghoff.github.io/rdev/)
- [Four ways to write assertion checks in R](https://blog.djnavarro.net/posts/2023-08-08_being-assertive/) -
  an article on four methods for writing assertions in R:
    - Use `stopifnot()` from base R
    - Use the [`assertthat`](https://github.com/hadley/assertthat) package (last updated March 2019)
    - Use the [`assertive`](https://bitbucket.org/richierocks/assertive/src/master/) package
      (last updated July 2020)
    - Use the [`assertr`](https://docs.ropensci.org/assertr/) package for data assertions, which
      works especially well for assertion checks against data frames used in data analysis
- I also revisited the [`validate`](https://github.com/data-cleaning/validate) package, a data
  validation rules engine, which includes the Data Validation Cookbook (in [Future Reading])

## rstats.wtf

[What They Forgot to Teach You About R](https://rstats.wtf) - a book based on a workshop taught by
current and former posit employees intended for self-taught data analysts. It's short, still in
draft, and offers good advice on R development (I use *almost* all the suggestions):

- Use an IDE (RStudio)
- Don't save `.RData`
- Restart R frequently
- Automate long workflows using [targets](https://docs.ropensci.org/targets/)
- Don't use `setwd()`
- Use RStudio projects
- Use [here](https://here.r-lib.org) and [fs](https://fs.r-lib.org) for paths
- Use "-" as space and "_" as a field delimiter in
  [file names](https://speakerdeck.com/jennybc/how-to-name-files)
- Use ISO standard dates (ISO 8601, YYYY-MM-DD)
- Use version control (git, GitHub)
- Use [renv](https://rstudio.github.io/renv/index.html), [rig](https://github.com/r-lib/rig/), and
  [homebrew](https://brew.sh)
- Use `.Renviron` and `.Rprofile`
- Use CRAN R and CRAN binaries to speed up package installation
- Don't use Conda (for Python)
- Update packages!
- Use [debugging](https://rstats.wtf/debugging-r) commands and RStudio for troubleshooting
- Search for error messages and read the source code to understand your problem

The book also suggests use of the [Posit Public Package Manager](https://p3m.dev/) (P3M). I'd *like*
to agree with this advice, however, I've had problems with their binaries; this package (rtraining)
doesn't work when using P3M, and works fine with CRAN binary packages.

## do4ds

[DevOps for Data Science](https://do4ds.com), another book written by a Posit employee was brought
to my attention thanks to a LinkedIn Post by [\@Corey Neskey](https://www.linkedin.com/in/cneskey/)
(It included a reference to rstats.wtf, which I read first). The book is split into 3 sections:

1. [DevOps Lessons for Data Science](https://do4ds.com/chapters/sec1/1-0-sec-intro.html)
   (Chapters 1-6)
1. [IT/Admin for Data Science](https://do4ds.com/chapters/sec2/2-0-sec-intro.html) (Chapters 7-14)
1. [Enterprise-Grade Data Science](https://do4ds.com/chapters/sec3/3-0-sec-intro.html)
   (Chapters 15-18)

I chose not to do the exercises in the book.

Chapters 1-6 cover basic software engineering concepts including library/package management, the
three-tier application model, types of projects (jobs, apps, reports, and APIs), connecting to
databases and APIs, logging and monitoring, deploying code with CI/CD in dev, test, and production
(including branching strategy) and Docker.

Useful references from the first section include:

- [pins](https://pins.rstudio.com), a package for publishing data, models, and other R objects to
  folders or cloud storage, including AWS, Azure, GCP, Google Drive, OneDrive, and Dropbox.
- [vetiver](https://vetiver.posit.co), R and Python libraries designed to version, deploy, and
  monitor a trained ML model.
- [Mastering Shiny](https://mastering-shiny.org/index.html), Hadley Wickham's book on R
  [Shiny](https://shiny.posit.co), published in 2021.

Chapters 7-14 cover basic infrastructure engineering concepts, including cloud computing (IaaS,
PaaS, and SaaS), the command line (terminals, shells, and SSH), Linux (Unix), scaling hardware,
networking, DNS, and TLS (HTTPS).

I disagreed with the recommendations on macOS terminals and configuration management in chapter 8; I
use macOS Terminal and `.zprofile` and `.zshrc` respectively. Given my background in infrastructure
and systems engineering, most of the section was not new to me, except for AWS, which I haven't
worked with.

There was one useful reference from the second section:

- [paws](https://github.com/paws-r/paws), an Amazon Web Services SDK for R.

Chapters 15-18 cover issues that come up in larger, "enterprise", organizations. It includes
discussion of networking and network security (including proxies), authentication and authorization
(including identity and access management: LDAP, SSO, MFA, RBAC and ABAC), infrastructure as code,
Docker and Kubernetes, and how enterprise policies affect use of R and Python packages.

The third section doesn't contain labs, and has one useful reference:

- [httr2](https://httr2.r-lib.org), tools for creating and modifying HTTP requests, then
  performing them and processing the results.

# Future Reading

R books on my reading list.

- [The R Manuals](https://rstudio.github.io/r-manuals/) - a re-styled version of the original R
  manuals, published using Quarto (starting with
  [*Writing R Extensions*](https://rstudio.github.io/r-manuals/r-exts/))
- [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/) - written by the author
  of [knitr](https://yihui.org/knitr/)
- [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/) - the follow-up to *The
  Definitive Guide*
- [The Data Validation Cookbook](https://data-cleaning.github.io/validate/) - a book on the R
  [validate](https://CRAN.R-project.org/package=validate) package
- [Discovering Statistics Using R](https://us.sagepub.com/en-us/nam/discovering-statistics-using-r/book236067) - recommended to me
  as an introduction to statistics using R
- While it's not an R book, I am interested in using
  [DiagrammeR](https://rich-iannone.github.io/DiagrammeR/) for graph visualizations (it also
  supports [Graphviz](https://graphviz.org) and [Mermaid](https://mermaid.js.org))

# R Dialects

An explanatory note on the R dialects of base R and tidyverse R.

The [R programming language](https://en.wikipedia.org/wiki/R_(programming_language)) is over 30
years old and has a large number of [packages](https://en.wikipedia.org/wiki/R_package) (R
libraries) that extend R. Unlike
[python](https://en.wikipedia.org/wiki/Python_(programming_language)) (a general purpose language),
R was designed specifically for analysis, visualization, and statistical modeling, which is why I
[chose R](https://www.information-safety.org/2020/09/11/working-with-r/) for data analysis: it has
built-in support for data structures like data frames (implemented in python using
[pandas](https://en.wikipedia.org/wiki/Pandas_(software))), vectors, packages for just about any
statistical tool you'd need, and of course, [ggplot2](https://en.wikipedia.org/wiki/Ggplot2). In
fairness, python is more popular, more robust, and a better tool for some tasks, like data
acquisition and machine learning (which were not priorities for my use).

Like many human languages, R has developed two distinct dialects: base R and tidyverse R. Base R
consists of the [packages](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-in-R)
included in the R distribution
*(`r paste(rownames(utils::installed.packages(priority = "base")), collapse = ", ")`)*, and the
[Tidyverse](https://en.wikipedia.org/wiki/Tidyverse) is a collection of packages that implement a
[domain-specific language](https://en.wikipedia.org/wiki/Domain-specific_language) for data
analysis, originally created by [Hadley Wickham](https://en.wikipedia.org/wiki/Hadley_Wickham).

In my experience, tidyverse R is better for data analysis, where base R is better for writing
packages - tidyverse functions are closer to natural language, but have many more dependencies.
[Comparing](https://atrebas.github.io/post/2019-03-03-datatable-dplyr/) two popular tools for data
manipulation, [dplyr](https://dplyr.tidyverse.org) (tidyverse R) and
[data.table](https://rdatatable.gitlab.io/data.table/) (base R) shows these differences.

This code snippet is from a short analysis of survey responses using dplyr:

```r
survey_results <- survey_import |>
  mutate(across(Q1:Q7, ~ case_when(
    .x == "strongly disagree" ~ 1,
    .x == "disagree" ~ 2,
    .x == "neither agree nor disagree" ~ 3,
    .x == "agree" ~ 4,
    .x == "strongly agree" ~ 5
  ))) |>
  mutate(Q8 = as.numeric(Q8 == "Yes")) |>
  arrange(end_date)
```

The code is reasonably easy to understand, even if you're not familiar with R.

The same code written in data.table isn't as clear:

```r
likert_5 <- c(
  "strongly disagree", "disagree", "neither agree nor disagree",
  "agree", "strongly agree"
)
q_likert <- paste0("Q", 1:7)
q_yesno <- "Q8"

survey_results <- copy(survey_import)
survey_results <- survey_results[
  , (q_likert) := lapply(.SD, \(x) as.numeric(factor(x, levels = likert_5))),
  .SDcols = q_likert
][
  , (q_yesno) := lapply(.SD, \(x) as.numeric(x == "Yes")),
  .SDcols = q_yesno
][
  order(end_date)
]
```

While it may be harder to read, data.table has some clear advantages: it is quite
[fast](https://h2oai.github.io/db-benchmark/), especially with very large datasets, and has no
[dependencies](https://github.com/Rdatatable/data.table/blob/master/DESCRIPTION) other than base R,
where dplyr has many.

These tradeoffs are why I tend to use tidyverse R for analysis and base R for functions (most
tidyverse expressions have functional equivalents in base R). Code used in data analysis should be
clear and easy to read, which tidyverse R excels at. Packaged functions provide documentation and
the source code isn't typically read, but many dependencies can be problematic; `R CMD check` will
raise a NOTE if there are too many imports.
