---
title: "R Training Log"
date: 2020-12-02
output:
  html_notebook:
    toc: TRUE
    toc_float:
      collapsed: FALSE
      smooth_scroll: FALSE
---

Notes on learning R and RStudio.

# Libraries

Load libraries and suppress warning and error messages. Thanks to this tidyverse [article](https://reprex.tidyverse.org/articles/articles/suppress-startup-messages.html).

```{r startup, message = FALSE, warning = FALSE}
library(tidyverse)
```

# Files

Files in this project:

- `.Rbuildignore`:
- `.Rhistory`:
- `.Rprofile`:
- `.Rproj.user`:
- `.git`:
- `.gitignore`:
- `.lintr`: [lintr](https://github.com/jimhester/lintr) configuration file
- `DESCRIPTION`:
- `LICENSE`:
- `LICENSE.md`:
- `NAMESPACE`:
- `README.Rmd`:
- `README.md`:
- `man`:
- `r-training-log.Rmd`:
- `r-training-log.nb.html`:
- `renv`:
- `renv.lock`:
- `rtraining.Rproj`:
- `setup-r`:
- `tests`:

# Not R

Not R, but has advice on git best practices and a guide to fixing commit mistakes: <https://sethrobertson.github.io>

# Learning R

Resources for learning R.

**Already using:** these are packages I'm already using in R. *Note:* much of this has been influenced by the work of [Hadley Wickham](http://hadley.nz/), who seems to be one of the (if not the most) significant contributors to the modern R and RStudio environment. Most of these have good integration with RStudio.

- [renv](https://github.com/rstudio/renv) - for package version management by project
- [styler](https://styler.r-lib.org/) and [lintr](https://github.com/jimhester/lintr) - for implementing the [tidyverse style guide](https://style.tidyverse.org/)
- Various [tidyverse](https://www.tidyverse.org/) packages, including:
  - dplyr
  - ggplot2
  - glue
  - lubridate
  - readr
  - stringr
  - tibble
  - tidyr
- Other packages:
  - gt
  - kableExtra
  - miniUI (for RStudio plugins)
  - openxlsx
  - knitr: there is a useful [knitr-spin.R](https://github.com/yihui/knitr/blob/master/inst/examples/knitr-spin.R) demo script that shows how to knit an R script (from <https://yihui.org/knitr/demo/stitch/>) - I briefly explored this as an alternative to notebooks
- R Notebooks leveraging GitHub Pages

When in doubt, start with the tidyverse, or RStudio sponsored projects, or built-in R.

**Completed:** I've read these.

- [R for Data Science](https://r4ds.had.co.nz/) - Hadley's intro book

**In progress:** I'm reading these now.

- [R Packages](https://r-pkgs.org/) - even more Hadley
- [testthat](https://testthat.r-lib.org/) - R testing framework
- [roxygen2](https://github.com/r-lib/roxygen2) - document functions inline

**Future:** I'm planning on adopting these to improve my code quality.

- Full [ggplot2](https://ggplot2.tidyverse.org/reference/index.html) documentation OR
- ggplot2 book, 3rd edition draft: <https://ggplot2-book.org>
- [Advanced R](https://adv-r.hadley.nz/) - more Hadley

**Also helpful:** some additional helpful links, either from me or recommended by others.

- [RStudio Education](https://education.rstudio.com/learn/beginner/) - the Beginners page
- [janitor](https://garthtarr.github.io/meatR/janitor.html) - automate data cleanup
- [purrr](https://purrr.tidyverse.org/) - for list iteration
- [broom](https://github.com/tidymodels/broom) - for getting classic stats objects into tidy shapes that are easier to work with

**References:** other good reference sites.

- [Yan Holtz](https://www.yan-holtz.com) has some good sites for data visualization:
  - [the R Graph Gallery](https://www.r-graph-gallery.com) - charts made with R and ggplot2
  - [from Data to Viz](https://www.data-to-viz.com) - recommendations on data visualization
  - He also has some tips on [R Markdown](https://holtzy.github.io/Pimp-my-rmd/)
- [MIT Roxygen Quick Reference](https://stuff.mit.edu/afs/athena/software/r/current/RStudio/resources/roxygen_help.html)

## R for Data Science

Selected highlights from the book:

### Patterns and models

[7.6 Patterns and models](https://r4ds.had.co.nz/exploratory-data-analysis.html#patterns-and-models) covers models and residuals, here showing how to reveal the expected relationship between cut and price:

```{r models}
library(modelr)

mod <- lm(log(price) ~ log(carat), data = diamonds)

diamonds2 <- diamonds %>%
  add_residuals(mod) %>%
  mutate(resid = exp(resid))

ggplot(data = diamonds2) +
  geom_point(mapping = aes(x = carat, y = resid))

ggplot(data = diamonds2) +
  geom_boxplot(mapping = aes(x = cut, y = resid))
```

### Non-Tidy Data

Hadley advocates for "tidy" data, but there are times when data is usefully [non-tidy](https://simplystatistics.org/2016/02/17/non-tidy-data/).

### Graphics: Learning more

From [28.8 Learning more](https://r4ds.had.co.nz/graphics-for-communication.html#learning-more-4):

"The absolute best place to learn more is the ggplot2 book: [_ggplot2: Elegant graphics for data analysis_](https://amzn.com/331924275X). It goes into much more depth about the underlying theory, and has many more examples of how to combine the individual pieces to solve practical problems. Unfortunately, the book is not available online for free, although you can find the source code at <https://github.com/hadley/ggplot2-book>.
 
Another great resource is the ggplot2 extensions gallery <https://exts.ggplot2.tidyverse.org/gallery/>. This site lists many of the packages that extend ggplot2 with new geoms and scales. It's a great place to start if you're trying to do something that seems hard with ggplot2."

### R Markdown Websites

From [29.8 Websites](https://r4ds.had.co.nz/r-markdown-formats.html#websites):

"With a little additional infrastructure you can use R Markdown to generate a complete website:

- Put your `.Rmd` files in a single directory. `index.Rmd` will become the home page.
- Add a YAML file named `_site.yml` provides the navigation for the site. For example:

```
name: "my-website"
navbar:
  title: "My Website"
  left:
    - text: "Home"
      href: index.html
    - text: "Viridis Colors"
      href: 1-example.html
    - text: "Terrain Colors"
      href: 3-inline.html
```

Execute `rmarkdown::render_site()` to build `_site`, a directory of files ready to deploy as a standalone static website, or if you use an RStudio Project for your website directory. RStudio will add a Build tab to the IDE that you can use to build and preview your site. 

Read more at <http://rmarkdown.rstudio.com/rmarkdown_websites.html>."

### R Notebook Advice

Hadley has some good suggestions for R Notebooks in [30 R Markdown workflow](https://r4ds.had.co.nz/r-markdown-workflow.html).

## R Packages

Notes from the book. I've started writing all code in the package structure described in the book and used by Hadley and the tidyverse. Doing so makes it easier to write tests, documentation, track dependencies, and provides convenient continous integration through `R CMD check`.

### System Setup

[Chapter 3](https://r-pkgs.org/setup.html) has good coverage of setting up a system for R development, I've incorporated what I learned from this into my "R Setup Log".

### Naming Things

Package names are pretty restricted, but Hadley has some good recommendation on [naming](https://r-pkgs.org/workflows101.html#naming) them, including the "[available](https://github.com/r-lib/available)" package, which I've started using.

### RStudio

There are some helpful tips in the [RStudio](https://r-pkgs.org/workflows101.html#benefits-of-rstudio-projects) section, including 2 key shortcuts to remember:

- control-. to "Go to file/function"
- alt-shift-k for the keyboard shortcut quick reference
- not a shortcut, but use [`proj_sitrep()`](https://usethis.r-lib.org/reference/proj_sitrep.html) to query active projects and working directory

### Writing Better Code

Chapter 6, [The Package Within](https://r-pkgs.org/package-within.html), highlights some sub-optimal R code and shows how to clean it up. It never occurred to me to use a [left join](https://r-pkgs.org/package-within.html#bravo-a-better-script-that-works) to clean up data, and I've already started using the technique.
